---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# checkrpkgs
`checkrpkgs` provides information on installing and configuring
[R](https://www.r-project.org/),
[RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
[Rtools](https://cran.r-project.org/bin/windows/Rtools/); obtaining information
about to-be-installed and already-installed packages; and how to get the source
code of R-functions.

## Installation
You can install `checkrpkgs` from
[GitHub](https://github.com/JesseAlderliesten/checkrpkgs) with:

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", quiet = FALSE)
}
remotes::install_github(repo = "JesseAlderliesten/checkrpkgs", dependencies = TRUE,
                        upgrade = FALSE, force = FALSE, quiet = FALSE,
                        build_vignettes = TRUE, lib = NULL,
                        verbose = getOption("verbose"))
```


## Example

The main content of `checkrpkgs` consists of vignettes:

- *Installing R, Rtools and RStudio*:
  `vignette("install_r", package = "checkrpkgs")`. This vignette contains
  instructions on installing and configuring [R](https://www.r-project.org/),
  [RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/).
- *Instructions about R packages*: `vignette("r_pkgs", package = "checkrpkgs")`.
  This vignette contains information on installing and getting information about
  R packages. It also explains how to obtain the source code of R functions.
- *Using Git and GitHub*: `vignette("git_github", package = "checkrpkgs")`. This
  vignette contains information about setting up and using Git and
  [GitHub](https://github.com/).


In addition, `checkrpkgs` contains the functions `find_nonfunc_pkgs()` and
`get_details_pkgs()`:

```{r example}
library(checkrpkgs)
find_nonfunc_pkgs(pkgs = "abcdef1") # package is absent

get_details_pkgs(pkgs = c("utils", "checkrpkgs")) # Details about packages.
```
