---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!-- badges: start -->
![](https://img.shields.io/github/r-package/v/JesseAlderliesten/checkrpkgs?color=blue)
[![R-CMD-check](https://github.com/JesseAlderliesten/checkrpkgs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JesseAlderliesten/checkrpkgs/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


# checkrpkgs
`checkrpkgs` provides information on installing and configuring
[R](https://www.r-project.org/),
[RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
[Rtools](https://cran.r-project.org/bin/windows/Rtools/); obtaining information
about to-be-installed and already-installed packages; and how to get the source
code of R functions.

## Installation
Visit the [checkrpkgs website](https://jessealderliesten.github.io/checkrpkgs/)
to explore the package., or install `checkrpkgs` from
[GitHub](https://github.com/JesseAlderliesten/checkrpkgs) using the following R
code (you need to run R as administrator):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", quiet = FALSE)
}
remotes::install_github(repo = "JesseAlderliesten/checkrpkgs",
                        dependencies = NA, upgrade = FALSE, force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = NULL,
                        verbose = getOption("verbose"))
```

## Example

The main content of `checkrpkgs` consists of vignettes:

- [Installing R, Rtools and RStudio](https://jessealderliesten.github.io/checkrpkgs/articles/install_r.html):
  `vignette("install_r", package = "checkrpkgs")`. This vignette contains
  instructions on installing and configuring [R](https://www.r-project.org/),
  [RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/).
- [R packages](https://jessealderliesten.github.io/checkrpkgs/articles/r_pkgs.html):
  `vignette("r_pkgs", package = "checkrpkgs")`.
  This vignette contains information on installing and getting information about
  R packages. It also explains how to obtain the source code of R functions.
- [Using Git and GitHub](https://jessealderliesten.github.io/checkrpkgs/articles/git_github.html):
  `vignette("git_github", package = "checkrpkgs")`. This
  vignette contains information about setting up and using Git and
  [GitHub](https://github.com/).


In addition, `checkrpkgs` contains the functions `check_pkgs()` and
`get_details_pkgs()`:

```{r example}
library(checkrpkgs)
check_pkgs(pkgs = "abcdef1") # package is absent

get_details_pkgs(pkgs = c("utils", "checkrpkgs"))
```

# Similar resources
- The book [What They Forgot to Teach You About R](https://rstats.wtf/) by
  Jennifer Bryan, Jim Hester, Shannon Pileggi, and E. David Aja
- The book [An introduction to R](https://intro2r.com/) by Alex Douglas, Deon
  Roos, Francesca Mancini, Ana Couto and David Lusseau

Developers might be interested in my package
[develcoder](https://jessealderliesten.github.io/develcoder/) with code to
develop R packages.
