---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

<!-- badges: start -->
![](https://img.shields.io/github/r-package/v/JesseAlderliesten/checkrpkgs?color=blue)
[![R-CMD-check](https://github.com/JesseAlderliesten/checkrpkgs/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/JesseAlderliesten/checkrpkgs/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


# checkrpkgs
`checkrpkgs` explains how to install and configure
[R](https://www.r-project.org/),
[RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
[Rtools](https://cran.r-project.org/bin/windows/Rtools/); how to install R
packages; how to get information about to-be-installed and already-installed
packages; how to get the source code of R functions; and how to set up and use
Git and GitHub with RStudio.

## Installation
Visit the [checkrpkgs website](https://jessealderliesten.github.io/checkrpkgs/)
to explore the package, or install `checkrpkgs` from
[GitHub](https://github.com/JesseAlderliesten/checkrpkgs) using the following R
code (you need to run R as administrator):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", quiet = FALSE)
}
remotes::install_github(repo = "JesseAlderliesten/checkrpkgs",
                        dependencies = NA, upgrade = FALSE, force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = NULL,
                        verbose = getOption("verbose"))
```

## Content
`checkrpkgs` consists of three vignettes:

- [Installing R, Rtools and RStudio](https://jessealderliesten.github.io/checkrpkgs/articles/install_r.html):
  `vignette("install_r", package = "checkrpkgs")`, explaining how to install and
  configure [R](https://www.r-project.org/),
  [RStudio](https://posit.co/products/open-source/rstudio/?sid=1), and
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/).
- [R packages](https://jessealderliesten.github.io/checkrpkgs/articles/r_pkgs.html):
  `vignette("r_pkgs", package = "checkrpkgs")`, explaining how to get
  information about to-be-installed and already-installed package and how to get
  the source code of R functions.
- [Using Git and GitHub](https://jessealderliesten.github.io/checkrpkgs/articles/git_github.html):
  `vignette("git_github", package = "checkrpkgs")`, explaining how to set up and
  use Git and [GitHub](https://github.com/) with RStudio.

# Similar resources
- The book [What They Forgot to Teach You About R](https://rstats.wtf/) by
  Jennifer Bryan, Jim Hester, Shannon Pileggi, and E. David Aja
- The book [An introduction to R](https://intro2r.com/) by Alex Douglas, Deon
  Roos, Francesca Mancini, Ana Couto and David Lusseau

Developers might be interested in my package
[develcoder](https://jessealderliesten.github.io/develcoder/) with code to
develop R packages.
