---
title: "Using Git and GitHub"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Using Git and GitHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction and notation
This vignette explains how to set up and use Git (a version control system) and
[GitHub](https://github.com/) (a website where files, usually code, can be
stored) with [RStudio](https://posit.co/products/open-source/rstudio).

Angled brackets (`<...>`) indicate place holders that should be replaced with
specific text to get working code or working file paths. For example,
`<username>` and `<repository>` are place holders that refer to a username and
repository name, e.g., `https://github.com/<username>/<repository>` could refer
to the URL
[https://github.com/JesseAlderliesten/checkrpkgs](https://github.com/JesseAlderliesten/checkrpkgs), the GitHub repository of this package.

Files on your PC are called often called 'local files', whereas files on GitHub
are called 'remote files'. Similarly, a folder on your PC (e.g.,
`C:\Program Files\R\R-4.6.0\library\checkrpkgs`) is often called a 'directory',
whereas a folder on GitHub (e.g., 
[https://github.com/JesseAlderliesten/checkrpkgs](https://github.com/JesseAlderliesten/checkrpkgs))
is often called a 'repository'.

## Setting up Git and GitHub
Create a [GitHub](https://github.com/) account, download a
[GitHub client](https://git-scm.com/tools/guis) such as
[`Git SCM`](https://git-scm.com/downloads/) or
[`Git for Windows`](https://gitforwindows.org/) and install it. To update Git for
Windows if you already have it, type `git\update-git-for-windows` in the
[shell](https://happygitwithr.com/shell).

To use Git and GitHub from [RStudio](https://posit.co/products/open-source/rstudio),
in RStudio go to `Tools` > `Global Options` >
`Git/SVN` and tick `Enable version control interface for RStudio projects`. The
`Git executable` field should point to the `git.exe` file, which probably is at
`C:\Program Files\Git\bin\git.exe` or in a hidden folder at
`C:\Users\<owner>\AppData\Local\Git\bin\git.exe` (in
Windows notation; outside Windows, these paths are written as
`C:/Program Files/Git/bin/git.exe` and
`C:/Users/<owner>/AppData/Local/Git/bin/git.exe`, respectively, see
[Paths in the shell]).
If the content of the field `Git executable` is not correct, open the `Git Bash`
[shell](https://happygitwithr.com/shell) (which was installed when installing
[`Git for Windows`](https://gitforwindows.org/)) by searching for `Git Bash` in
Windows' `Start` menu. In the Git Bash shell, run `where git.exe`
to get the location of the Git executable.

To associate Git with your GitHub account, you need to provide your name
(this name is listed in GitHub with the changes you make and does **not** have to
be your GitHub username), and the email associated with your GitHub account.
See the Chapter
[`Introduce yourself to Git`](https://happygitwithr.com/hello-git.html) from
[`Happy Git and GitHub for the useR`](https://happygitwithr.com/) for more details.

```
git config --global user.name '<Jane Doe>'
git config --global user.email '<jane@example.com>'
git config --global --list
```
The last line should return the username and email address you just entered.

### Using personal access tokens (PATs)
On the use of personal access tokens (PATs) instead of a username and password,
see:

- GitHub [documentation](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens)
- The chapter about [`Personal Access Tokens`](https://happygitwithr.com/https-pat.html)
  in [`Happy Git and GitHub for the useR`](https://happygitwithr.com/)
- A [vignette](https://usethis.r-lib.org/articles/git-credentials.html) from
  package [`usethis`](https://CRAN.R-project.org/package=usethis)


## Using Git and GitHub
To work with GitHub in RStudio, you should open the R Project file (i.e., a file
with extension `.Rproj`), **not** the R script (i.e., a file with extension
`.R`). Then the `Git` menu will be visible as a tab in the
[`Environment pane`](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html).

`Pull` to get changes from GitHub incorporated in
your PC, and handle any conflicts to get
directory on your PC up-to-date with the repository on GitHub:

- in RStudio: use the `pull` button (downward arrow) in the `Git` menu
- or in the [shell](https://happygitwithr.com/shell):
  `git pull https://github.com/<username>/<repository>`

Save the modified R file after you have made some changes, only then will the
name of the file appear in the `Git` menu of RStudio to review changes:

- in RStudio: check the `Staged` box in front of the relevant filename, use the
  `Diff` button in the `Git` menu to get an overview of the changes to the file,
  in the box `Commit message` you should describe the changes and why you made
  them, and use the `Commit` button.
  `Pull` again (downward arrow) to make sure the directory on your PC is
  up-to-date, and
  handle any conflicts. Then `push` (upward arrow) to incorporate the changes in
  the repository on GitHub.
- or in the [shell](https://happygitwithr.com/shell): compare the content of two
  files, see the instructions in the section [Comparing files] below. Next, use
  `git commit -m '<your commit message>' <path>/<and>/<filename>.R`
  `git push https://github.com/<username>/<repository>` to commit changes.

If a `Push` leads to an error because of an invalid username or password, `Push`
again, then you will be asked for a personal access token (PAT; see the section
[Using personal access tokens (PATs)] above). After you have entered the PAT once,
RStudio will remember it.

To `Push` only some of the changes you made, use the `Diff` button in the `Git`
menu and select the line or lines of code containing the changes you want to
`Push`. Then a button `Stage line` or `Stage chunck` will show up next to the
selected code. Click that button to `Push` only the selected lines.

## Adding a new file
To add a file to GitHub that is not there yet, first add it to the R project
folder on your PC (see the output of `getwd()`), then check the `Staged` box in
front of the relevant filename in the `GitHub` pane of RStudio and commit it as
described [above][Using Git and GitHub].

Using the [shell](https://happygitwithr.com/shell) to add a new file is more
involved: if the working directory **of the shell** (see the output of `pwd`) is
**not** the folder where the to-be-added file is in, the working directory has
to be changed to that folder (e.g., using `cd <path/to/folder>` in the shell),
or the path has to be added in front of the filename (dragging the file onto the
shell will copy the path to the shell).

Then let `git` know the new file is there by typing (it is convenient to use
`tab`-completion to select files): `git add <filename>.<extension>`.

## Comparing files

### Using GitHub
- For a chronological overview of all commits (i.e., across all files) in a
  particular branch of a repository, use an URL of the form
  `http://github.com/<username>/<repository>/commits`, or go to the main page of the
  repository and in the `Code` panel click the `clock` icon at the top of the
  file overview.
- To compare different branches of a repository, use three dots between their
  names, giving an URL of the form
  `https://github.com/<username>/<repository>/compare/<branch>...<otherbranch>`,
  for example
  `https://github.com/JesseAlderliesten/checkrpkgs/compare/main...devel` that
  compares the `main` and `devel` branches of this package. Details:
  [here](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-comparing-branches-in-pull-requests#three-dot-and-two-dot-git-diff-comparisons)
- To compare different commits in a repository, use two dots between their
  commit IDs (the shortened SHA codes displayed at the right of each commit in
  the commit history, and at the top of the commit), giving an URL of the form
  `https://github.com/<username>/<repository>/compare/<ID_commit>...<ID_othercommit>`,
  for example
  `https://github.com/JesseAlderliesten/checkrpkgs/compare/e8eb0a1...e60a155`
  that compares two commits. Details:
  [here](https://docs.github.com/en/pull-requests/committing-changes-to-your-project/viewing-and-comparing-commits/comparing-commits)
- See also section [Repositories: download, fork, or clone?] below.

### Not using GitHub
R scripts can be compared using `tools::Rdiff()`: use the quoted file paths
(i.e., the directories and file names, **including** the extensions) of the two
files as arguments `from` and `to`:  
`tools::Rdiff(from = "<path/and/filename_file1>.R", to = "<path/and/filename_file2>.R")`

For a nicer output, compare R scripts using the `Bash`
[shell](https://happygitwithr.com/shell): open the `Bash` shell and copy the
file paths (i.e., the directories and file names, **including** the extensions)
of the two files to be compared into the shell on the same line, and press `Enter`:
`git diff --no-index '<path/and/filename_file1>.R' '<path/and/filename_file2>.R'`

Notes:

- `--no-index` makes it possible to compare files that are **not** under version
  control in Git.
- Using quotes (`''`) around the paths ensures this also works when they contain
  spaces.
- The two paths should be on the same line, i.e., **not** separated by a newline.
- Although the scroll bar in `Git Bash` seems to indicate the end of the file is
  reached (and scrolling with the mouse does not work), usually a colon (`:`)
  will be displayed left to the cursor to indicate that only part of the file is
  shown. Use the `down arrow` key to see the whole file until the end of the
  file is reached, which is indicated by `(END)`.
- The location of a change is indicated at the top of a changed chunk. E.g.,
  `-13,5 +14,9` indicates a deletion at character five of line thirteen and an
  insertion at character nine of line fourteen.


## Deleting a file
To delete a file, delete the file from your PC, then `commit`
the change (i.e., deletion of a file), using the `commit message` to describe
why the file was deleted. Then `pull` to make sure directory on your PC is up to
date with the repository on GitHub, and then `push` the change to the GitHub
repository.

Alternatively, open the file in the GitHub repository, click the three dots
at the top-right of the file > `delete` >
`commit`. Use the commit message to describe why the file was deleted. After
that, find the file on your PC and delete it. Then `pull`.

Deleted files and their history can still be viewed on GitHub, e.g., by
[finding the commit][Using GitHub] in which the file was deleted.


## Moving or renaming a file
See the GitHub documentation
[here](https://docs.github.com/en/repositories/working-with-files/managing-files/moving-a-file-to-a-new-location)
and
[here](https://docs.github.com/en/repositories/working-with-files/managing-files/renaming-a-file).


## Repositories: download, fork, or clone?
There are several ways to copy code from a GitHub repository to your PC:

- To be able to `push` your changes back to a GitHub repository to which you
  do **not** have writing access, you need to `fork` the repository: use the
  `fork` button on the GitHub page of the repository and then use
  `create a new fork`. This creates a copy of the repository in
  your own GitHub repository. Next, you have to `clone` your copy to your PC,
  see the next point.
- To be able to push your changes back to a GitHub repository to which you
  **do** have writing access (e.g., to work on a fork you created in the step
  above; or to work on your project from another PC), you have to clone the GitHub
  repository to your PC: use the green `Code` button on the GitHub page of the
  repository (if you
  forked a repository, you need the `Code` button of **your fork**, not of the
  original repository), copy the
  URL to the clipboard (i.e., do **not** use `download ZIP`), create a new R
  project in RStudio (`File` > `New Project` > `Version control` > `Git`), paste
  the repository URL (`https://github.com/<username>/<repository>`) in the
  designated field, select the desired location on your PC, and create the
  project. The same repository URL can be used when using shell commands to
  clone a repository, either with its complete history by using
  `git clone https://github.com/<username>/<repository>` or with only the last
  commit by using `git clone --depth=1 https://github.com/<username>/<repository>`.
- To download code without being able to push your changes back to a GitHub
  repository, download the repository by using the green `Code` button on the
  GitHub page of the repository,
  choose `Download ZIP` and unzip the downloaded file (right-click on them and
  choose `extract all`). To be able to let R use the R files correctly, R has to
  know where the files are. If the downloaded files are an R package, move it to
  a location where R looks for packages (given by
  `cat(normalizePath(.libPaths()))`, something like
  `C:\Program Files\R\R-4.6.0\library` or
  `C:\Users\<owner>\AppData\Local\R\win-library\4.6`). Then open the `.Rproj`
  file that has the same name as the repository. If the downloaded files are
  not a package but an R project, open the `.Rproj` file to change the working
  directory to the folder where that project file is. If the downloaded files
  are not an R project, you have to move them to the working directory (see the
  output of `getwd()`), or use `setwd(<path/to/files>)` to change the working
  directory to the location where the files are.

### Installing a package from GitHub
The following code can be used to install packages from
[GitHub](https://github.com/) (for details see the section
`Installing packages` > `Github` in the vignette *R packages*:
`vignette("r_pkgs", package = "checkrpkgs")`):

```{r, eval = FALSE}
pkgs_new <- "JesseAlderliesten/checkrpkgs"
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_github(repo = grep(pattern = "/", x = pkgs_new, value = TRUE,
                                    fixed = TRUE),
                        dependencies = NA, upgrade = "ask", force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = .libPaths(),
                        verbose = getOption("verbose"))
```

### Further documentation
In addition to section [Using GitHub] above, see:

- Branches: GitHub documentation about
[`branches`](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-branches)
and the website [Learn Git Branching](https://learngitbranching.js.org/).

- Cloning (clones) and forking (forks) a repository: GitHub documentation about
  [`cloning`](https://docs.github.com/en/repositories/creating-and-managing-repositories/cloning-a-repository) and
  [`forks`](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/about-forks)
  and the section [`Fork and clone`](https://happygitwithr.com/fork-and-clone)
  from [`Happy Git and GitHub for the useR`](https://happygitwithr.com/)
- GitHub documentation about
  [`pull requests`](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request)


## GitHub Actions
See the section `Automate checks` in the vignette *Package setup* and section
`Use GitHub Actions` in the vignette *Package development*, both from package
`develcoder`: `vignette("pkg_setup", package = "develcoder")` and
`vignette("pkg_devel", package = "develcoder")`.


## Common shell commands
For an overview of [shell](https://happygitwithr.com/shell) commands, see the
documentation of [`Git SCM`](https://git-scm.com/docs), the section about the
[`shell`](https://happygitwithr.com/shell#basic-shell-commands) from 
[`Happy Git and GitHub for the useR`](https://happygitwithr.com/) (a summary of
which is given below), or in the `BASH shell` type `git config` or
`git help <command>` (replace `<command>` by the command you want help about).

- list files: `ls` (use `ls -a` to also show hidden files)
- list remote repositories: `git remote -v`
- show status of repositories: `git status`
- show user details: `git config --global --list`
- working directory: change it with `cd` (e.g.,
  `cd 'D:/Userdata/<owner>/Documents/GIT/<somefolder>'`); navigate to it:
  `cd ~`; print it: `pwd`

### Paths in the shell
When entering paths in the [shell](https://happygitwithr.com/shell), use the
forward slash (`/`) as file separator instead of the
Windows-default backslash (`\`). If the path you want to specify contains spaces
(e.g., `D:\Userdata\My Account\...`), you need to use quotes around the path (e.g.,
`"D:/Userdata/My Account/..."`). `Tab`-completion can be used when entering paths:
single `tab` to select an option, double `tab` to see multiple options. Dragging
a file into the shell gives the absolute path to that file. The current and
parent directory can be indicated by a single (`.`) or two (`..`) dots in file
paths, respectively. On paths and file separators in R, see the 'Notes on paths'
in `help("is_path", package = "checkinput")`.


## Documentation

Official documentation

- [Git](https://git-scm.com/docs)
- [Git: setting up](https://docs.github.com/en/get-started/git-basics/set-up-git),
- [GitHub: setting up](https://docs.github.com/en/get-started/onboarding/getting-started-with-your-github-account)
- [GitHub: general](https://docs.github.com/en)
- [GitHub: searching](https://docs.github.com/en/search-github)
- [GitHub: status](https://www.githubstatus.com/)
- [GitHub: workflow](https://docs.github.com/en/get-started/using-github/github-flow)

Books

- [`Happy Git and GitHub for the useR`](https://happygitwithr.com/) by Jennifer
  Bryan
- [`Pro Git Book`](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub

Chapters

- [`Version control`](https://docs.posit.co/ide/user/ide/guide/tools/version-control.html) from Posit's
  [`RStudio User Guide`](https://posit.co/products/open-source/rstudio)
- [`Collaboration`](https://epirhandbook.com/en/new_pages/collaboration.html) from
  [`The Epidemiologist R Handbook`](https://epirhandbook.com/en/)
