---
title: "Using Git and GitHub"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Using Git and GitHub}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


## Introduction and notation
This vignette explains how to set up and use Git (a version control system) and
[GitHub](https://github.com/) (a website where files, usually code, can be
stored) with [RStudio](https://posit.co/products/open-source/rstudio).

Files on your PC are called 'local', whereas files on GitHub are called 'remote'.
In addition, a folder on your PC (e.g.,
`C:\Program Files\R\R-4.6.0\library\checkrpkgs`) is usually called a
'directory', whereas a folder on GitHub (e.g., 
[https://github.com/JesseAlderliesten/checkrpkgs](https://github.com/JesseAlderliesten/checkrpkgs))
is usually called a 'repository'.

Text between angled brackets (`<...>`) is used to refer to
text that should be replaced with specific text to get working code or working
file paths. For example, `<username>` and `<repository>` are used as place
holders to refer to a username and repository name: `<username>` and
`<repository>` in `https://github.com/<username>/<repository>` should be
replaced with `JesseAlderliesten` and `checkrpkgs`, respectively, to obtain the
URL
[https://github.com/JesseAlderliesten/checkrpkgs](https://github.com/JesseAlderliesten/checkrpkgs)
referring to the GitHub repository of this package.


## Setting up Git and GitHub
Create a [GitHub](https://github.com/) account, download a
[GitHub client](https://git-scm.com/tools/guis) such as
[Git SCM](https://git-scm.com/downloads) or
[Git for Windows](https://gitforwindows.org/) and install it. To update Git for
Windows if you already have it, type `git\update-git-for-windows` in the
[shell](https://happygitwithr.com/shell).

To use Git and GitHub from [RStudio](https://posit.co/products/open-source/rstudio):
in RStudio at `Tools` > `Global Options` >
`Git/SVN`, tick `Enable version control interface for RStudio projects`. Run
`where git.exe` in the [shell](https://happygitwithr.com/shell) to get the
location of the Git executable: probably `C:\Program Files\Git\bin\git.exe` or
`C:\Users\<owner>\AppData\Local\Git\bin\git.exe` (usually a hidden folder) in
Windows notation; in R these paths are `C:/Program Files/Git/bin/git.exe` or
`C:/Users/<owner>/AppData/Local/Git/bin/git.exe`, see [Paths in the shell].

Open the `Git Bash` [shell](https://happygitwithr.com/shell) (which was
installed when installing [Git for Windows](https://gitforwindows.org/)) by
typing `Git Bash` in Windows' `Start` menu and press `Enter`.

To associate Git with your GitHub account, you need to provide your name
(this name is listed in GitHub with the changes you make and does **not** have to
be your GitHub username), and the email associated with your GitHub account.
See the Chapter
[Introduce yourself to Git](https://happygitwithr.com/hello-git.html) from
[Happy Git and GitHub for the useR](https://happygitwithr.com/) for more details.

```{Git Bash}
git config --global user.name '<Jane Doe>'
git config --global user.email '<jane@example.com>'
git config --global --list
```
The last line should return the username and email address you just entered.

### Using personal access tokens (PATs)
On the use of personal access tokens (PATs) instead of a username and password,
see:

- GitHub [documentation](https://docs.github.com/en/authentication/keeping-your-account-and-data-secure/managing-your-personal-access-tokens);
- The Chapter about [Personal Access Tokens](https://happygitwithr.com/https-pat.html)
  in [Happy Git and GitHub for the useR](https://happygitwithr.com/);
- A [vignette](https://usethis.r-lib.org/articles/git-credentials.html) from
  package [usethis](https://CRAN.R-project.org/package=usethis).


## Using Git and GitHub
To work with GitHub in RStudio, you should open the R Project file (i.e., a file
with extension `.Rproj`), **not** the R script (i.e., a file with extension
`.R`). Then the `Git` menu will be visible as a tab in the
[Environment pane](https://docs.posit.co/ide/user/ide/guide/ui/ui-panes.html).

`Pull` to get changes from the remote repository (i.e., `GitHub`) incorporated in
the local directory (i.e., your PC), and handle any conflicts to get the local
directory up-to-date with the remote repository:

- in RStudio: use the `pull` button (downward arrow) in the `Git` menu;
- in the [shell](https://happygitwithr.com/shell):
  `git pull https://github.com/<username>/<repository>`.

Save the modified R file after you have made some changes, only then will the
name of the file appear in the `Git` menu of RStudio to review changes:

- in RStudio: check the box in front of the relevant filename, use the
  `Diff` button in the `Git` menu to get an overview of the changes to the file,
  describe the changes in the box `Commit message`, and use the `Commit` button.
  `Pull` again (downward arrow) to make sure the local directory is up-to-date, and
  handle any conflicts. Then `push` (upward arrow) to incorporate the changes in
  the remote repository.
- in the [shell](https://happygitwithr.com/shell): to compare the content of two
  files, see the instructions in the section [Comparing files] below. Next, use
  `git commit -m '<your commit message goes here>' <path>/<and>/<filename>.R`
  `git push https://github.com/<username>/<repository>` to commit changes.

If a `Push` leads to an error because of an invalid username or password, `Push`
again, then you will be asked for a personal access token (PAT; see the section
[Using personal access tokens (PATs)] above). After you have entered the PAT once,
RStudio will remember this token for next times.


## Adding a new file
To add a file to GitHub that is not there yet, first add it to the local folder
(i.e., on your PC), then let `git` know it is there by typing (it is convenient
to use `tab`-completion to select files): `git add <filename>.<extension>`.

If the working directory is **not** the folder where the to-be-added file is in,
the working directory has to be set to that folder (e.g., using
`cd <path/to/folder>`, in the [shell](https://happygitwithr.com/shell) or the
path has to be added in front of the filename (dragging the file onto the shell
will copy the path to the shell).

## Comparing files

### Using GitHub
- For a chronological overview of all commits (i.e., across all files) in a
  particular branch of a repository, use an URL of the form
  `http://github.com/<username>/<repository>/commits`, or go to the main page of the
  repository and in the `Code` panel click the `clock` icon at the top of the
  file overview.
- Comparing different branches: see
  [here](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-comparing-branches-in-pull-requests#three-dot-and-two-dot-git-diff-comparisons)
- Comparing commits: see
  [here](https://docs.github.com/en/pull-requests/committing-changes-to-your-project/viewing-and-comparing-commits/comparing-commits)
- See also section [Repositories: download, fork, or clone?] below.

### Not using GitHub
R scripts can be compared using `tools::Rdiff()`: use the quoted file paths
(i.e., the directories and file names, **including** the extensions) of the two
files as arguments `from` and `to`:  
`tools::Rdiff(from = "<path/and/filename_file1>.R", to = "<path/and/filename_file2>.R")`

For a nicer output, compare R scripts using the `Bash`
[shell](https://happygitwithr.com/shell): open the `Bash` shell and copy the
file paths (i.e., the directories and file names, **including** the extensions)
of the two files to be compared into the shell on the same line, and press `Enter`:
`git diff --no-index '<path/and/filename_file1>.R' '<path/and/filename_file2>.R'`

Note:

- The flag `--no-index` makes it possible to compare files that are **not**
  under version control in `Git`.
- Using quotes (`''`) around the paths ensures this also works when they contain
  spaces.
- The two paths should be on the same line, i.e., **not** separated by a newline.
- Although the scroll bar in `Git Bash` seems to indicate the end of the file is
  reached (and scrolling with the mouse does not work), usually a colon (`:`)
  will be displayed left to the cursor to indicate that only part of the file is
  shown. Use the `down arrow` key to see the whole file until the end of the
  file is reached, which is indicated by `(END)`.
- The location of a change is indicated at the top of a changed chunk. E.g.,
  `-13,5 +14,9` indicates a deletion at character five of line thirteen and an
  insertion at character nine of line fourteen.


## Deleting a file
Consider moving files to a folder `archive` instead of deleting them. To delete
a file, delete the file from the local directory (i.e., your PC), then `commit`
the change (i.e., deletion of a file), using the `commit message` to describe
why the file was deleted. Then `pull` to make sure the local directory is up to
date with the remote repository, and then `push` the change to the remote
repository.

Alternatively, open the file on the remote repository (i.e., on the `GitHub`
website), click the three dots at the top-right of the file > `delete` >
`commit`. Use the commit message to describe why the file was deleted. After
that, find the file in the local directory and delete it. Then `pull`.

Deleted files and their history can still be viewed on GitHub, e.g., by finding
the commit in which the file was deleted.


## Moving or renaming a file
See the GitHub documentation
[here](https://docs.github.com/en/repositories/working-with-files/managing-files/moving-a-file-to-a-new-location)
and
[here](https://docs.github.com/en/repositories/working-with-files/managing-files/renaming-a-file).


## Repositories: download, fork, or clone?
There are several ways to get code from a `GitHub` repository to your PC:

- To be able to `push` your changes back to a `GitHub` repository to which you
  do **not** have writing access, you need to `fork` the repository: use the
  `fork` button > `create a new fork`. This creates a copy of the repository in
  your own GitHub repository. Next, you have to `clone` your copy to your PC,
  see the next point.
- To be able to push your changes back to a GitHub repository to which you
  **do** have writing access (e.g., to work on your project from another PC; or
  to work on a fork you created in the step above), you have to clone the GitHub
  repository to your PC: use the green `Code` button in the repository (if you
  forked a repository, you need the `Code` button of your fork, not of the
  original repository), copy the
  URL to the clipboard (i.e., do **not** use `download ZIP`), create a new R
  project in RStudio (`File` > `New Project` > `Version control` > `Git`), paste
  the repository URL (`https://github.com/<username>/<repository>`) in the
  designated field, select the desired location on your PC, and create the
  project. The same repository URL can be used when using shell commands to
  clone a repository, either with its complete history by using
  `git clone https://github.com/<username>/<repository>` or with only the last
  commit by using `git clone --depth=1 https://github.com/<username>/<repository>`.
- To download code without being able to push your changes back to a GitHub
  repository, you can download the repository by using the green `Code` button,
  choosing `Download ZIP` and unzip the downloaded file (right-click on them and
  choose `extract all`). To be able to let R use the package correctly, move the
  package to a location where R looks for packages (given by
  `cat(normalizePath(.libPaths()))`, something like
  `C:\Program Files\R\R-4.6.0\library` or
  `C:\Users\<owner>\AppData\Local\R\win-library\4.6`. Then open the `.Rproj`
  file (which has the same name as the repository).

### Installing a package from GitHub
The following code can be used to install packages from
[GitHub](https://github.com/) (for details see the section
'Installing packages > Github' in the vignette *R packages*:
`vignette("r_pkgs", package = "checkrpkgs")`):

```{r, eval = FALSE}
remotes::install_github(repo = pkgs_new[basename(pkgs_new) != pkgs_new],
                        dependencies = NA, upgrade = "ask", force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = .libPaths(),
                        verbose = getOption("verbose"))
```

### Further documentation
In addition to section [Using GitHub] above, see:

- Branches: GitHub documentation about [branches](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/about-branches) and https://learngitbranching.js.org/

- Cloning (clones) and forking (forks) a repository: GitHub documentation about
  [cloning](https://docs.github.com/en/repositories/creating-and-managing-repositories/cloning-a-repository) and
  [forks](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/working-with-forks/about-forks)
  and the section [Fork and clone](https://happygitwithr.com/fork-and-clone)
  from [Happy Git and GitHub for the useR](https://happygitwithr.com/);
- GitHub documentation about
  [pull requests](https://docs.github.com/en/pull-requests/collaborating-with-pull-requests/proposing-changes-to-your-work-with-pull-requests/creating-a-pull-request)


## GitHub Actions
See the section 'Automate checks' in the vignette *Package setup* and section
'Use GitHub Actions' in the vignette *Package development*, both from package
`develcoder`: `vignette("pkg_setup", package = "develcoder")` and
`vignette("pkg_devel", package = "develcoder")`.


## Common shell commands
For an overview of [shell](https://happygitwithr.com/shell) commands, see the
documentation of [Git SCM](https://git-scm.com/docs), the section about the
[shell](https://happygitwithr.com/shell#basic-shell-commands) from 
[Happy Git and GitHub for the useR](https://happygitwithr.com/) (a summary of
which is given below), or in the `BASH shell` type `git config` or
`git help <command>` (replace `<command>` by the command you want help about).

- list files: `ls` (use `ls -a` to also show hidden files)
- list remote repositories: `git remote -v`
- show status of repositories: `git status`
- show user details: `git config --global --list`
- working directory: change it with `cd` (e.g.,
  `cd 'D:/Userdata/<owner>/Documents/GIT/<somefolder>'`); navigate to it:
  `cd ~`; print it: `pwd`

### Paths in the shell
When entering paths in the [shell](https://happygitwithr.com/shell), use the
forward slash (`/`) as file separator instead of the
Windows-default backslash (`\`). If the path you want to specify contains spaces
(e.g., `D:/Userdata/My Account/...`), use quotes around the path (e.g.,
`"D:/Userdata/My Account/..."`). `Tab`-completion can be used when entering paths:
single `tab` to select an option, double `tab` to see multiple options. Dragging
a file into the shell gives the absolute path to that file. The current and
parent directory can be indicated by a single (`.`) or two (`..`) dots in file
paths, respectively. On paths and file separators in R, see the 'Notes on paths'
in `help("is_path", package = "checkinput")`.


## Documentation

Official documentation

- [Git](https://git-scm.com/docs)
- [Git: setting up](https://docs.github.com/en/get-started/git-basics/set-up-git),
- [GitHub: setting up](https://docs.github.com/en/get-started/onboarding/getting-started-with-your-github-account)
- [GitHub: general](https://docs.github.com/en)
- [GitHub: searching](https://docs.github.com/en/search-github)
- [GitHub: status](https://www.githubstatus.com/)
- [GitHub: workflow](https://docs.github.com/en/get-started/using-github/github-flow)

Books

- [Happy Git and GitHub for the useR](https://happygitwithr.com/) by Jennifer
  Bryan
- [Pro Git Book](https://git-scm.com/book/en/v2) by Scott Chacon and Ben Straub

Chapters

- [Version control](https://docs.posit.co/ide/user/ide/guide/tools/version-control.html) from Posit's
  [RStudio User Guide](https://posit.co/products/open-source/rstudio)
- [Collaboration](https://epirhandbook.com/en/new_pages/collaboration.html) from
  [The Epidemiologist R Handbook](https://epirhandbook.com/en/)

On git tags and releases:

- https://docs.github.com/en/repositories/releasing-projects-on-github/about-releases
- https://www.toolsqa.com/git/git-tags/
- https://www.jetbrains.com/help/dataspell/use-tags-to-mark-specific-commits.html
