---
title: "Instructions about R packages"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Instructions about R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction and notation
This vignette contains information on installing and getting information about R
packages. It also explains how to obtain the source code of R functions.

In this vignette, text between angled brackets (`<...>`) is used to refer to
text that should be replaced with specific text to get working code or working
file paths. For example, `<pkgname>` is used as a place holder to refer in
general to a package name: `<pkgname>` should be replaced with `utils` if you
want to obtain information about package `utils`, and with `methods` if you want
to obtain information about package `methods`. Similarly, `<funcname>` is a
place holder to refer to a function name that should be filled in with a
specific function name to get working code.

In this vignette, calls to functions are frequently written in the form
`<pkgname>::<funcname>()`. This notation is used to make clear which package is
used and, through the brackets, that a function is indicated. In normal code,
one would use `library(<pkgname>)` followed by `<funcname>()`. For example, in
this vignette the notation `utils::citation()` or
`utils::citation(package = "checkrpkgs")` is used to show how to cite R or the
package `checkrpkgs`, indicating that the function `citation` is defined in
package `utils`. In normal code, one would use `library(utils)` followed by
`utils::citation()` and `utils::citation(package = "checkrpkgs")`.

## R packages
An R
[package](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-is-the-difference-between-package-and-library_003f)
is a standardized collection of material extending R by providing code, data, or
documentation.

The
[base](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
packages are always installed together with R. The
[recommended](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-from-CRAN)
packages are installed with binary distributions of R. Together, the base and
recommended packages are the 'high priority' packages. Since R 4.4.0,
`tools::standard_package_names()` contains a list with the names of these
packages. To see which high-priority packages are currently installed, run
`sort(unname(installed.packages(priority = "high")[, "Package"]))`. Package
'translations' is not a recommended package, but will be installed if that
option is set during the installation of R.

### Installing packages
To install a package, run R or RStudio as administrator: right-click on the R or
RStudio icon and select `Run as administrator`. Packages can be obtained from
several websites, called 'repositories' (see below). After installing a package,
you need to run [library](../help/library)`(<pkgname>)` to be able to use the
functions of that package.

#### CRAN
The [Comprehensive R Archive Network](https://cran.r-project.org/) (CRAN) is the
main repository of R [packages](https://cran.r-project.org/web/packages/index.html).
The following code can be used to install packages from CRAN:

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
# Select packages from 'pkgs_new' that are not installed or not functional
pkgs_install <- pkgs_new[!vapply(X = pkgs_new, FUN = requireNamespace,
                                 FUN.VALUE = logical(1), quietly = TRUE)]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

CRAN has thematic package collections known as
[task views](https://cran.r-project.org/web/views/). To install all core packages
of a task view, install package [ctv](https://CRAN.R-project.org/package=ctv)
and run `ctv::install.views("<taskviewname>", coreOnly = TRUE)`. To update these
packages, use `ctv::update.views("<taskviewname>", coreOnly = TRUE)`.

Packages from CRAN that have been recently archived are available at
[CRANhaven](https://www.cranhaven.org/).

#### BioConductor
Each [Bioconductor](https://bioconductor.org/) version contains specific
versions of packages from [CRAN](https://cran.r-project.org/web/packages/index.html)
and [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
that are consistent with each other and with a
[specific version](http://bioconductor.org/about/release-announcements/#release-versions)
of R, preventing version conflicts between R packages. Bioconductor also has
thematic package collections known as
[BiocViews](https://bioconductor.org/packages/release/BiocViews.html).

The following code can be used to install packages from BioConductor release
[version](http://bioconductor.org/about/release-announcements/#release-versions)
3.22. This code installs the
[BiocManager](https://CRAN.R-project.org/package=BiocManager) package from
[CRAN](https://CRAN.R-project.org) that is then used to install packages from
`Bioconductor`:

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = pkgs_new, lib = .libPaths(), dependencies = NA,
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = FALSE, ask = TRUE, checkBuilt = TRUE,
                     force = FALSE, version = "3.22")
```

#### Github
The following code can be used to install packages from
[GitHub](https://github.com/): it installs the
[remotes](https://CRAN.R-project.org/package=remotes) package that is needed to
install packages from GitHub. `grep(pattern = "/", x = pkgs_new, value = TRUE)`
selects the elements of `pkgs_new` which contain a slash because for argument
`pkgs` of `remotes::install_github()`, each element of `pkgs` should contain the
author and repository name (e.g., `"JesseAlderliesten/checkrpkgs"`) or the full
URL to a package (e.g., `"https://github.com/JesseAlderliesten/checkrpkgs"`).

```{r, eval = FALSE}
pkgs_new <- c("JesseAlderliesten/checkrpkgs",
              "https://github.com/JesseAlderliesten/progutils")
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_github(repo = grep(pattern = "/", x = pkgs_new, value = TRUE),
                        dependencies = NA, upgrade = "ask", force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = .libPaths(),
                        verbose = getOption("verbose"))
```

#### Other repositories and mirrors ####
Examples of other repositories for R packages are:

- [Neuroconductor](https://neuroconductor.org/list-packages/all)
- [R-Forge](https://r-forge.r-project.org/) with a 
  [GitHub mirror](https://github.com/r-forge) and thematic package
  [collections](https://r-forge.r-project.org/softwaremap/trove_list.php) 
- [R-multiverse](https://r-multiverse.org/overview.html) with
  thematic [collections](https://r-multiverse.org/topics/)
- [rOpenSci](https://ropensci.org/packages/all/) with thematic package
  [collections](https://ropensci.org/packages/)
- [R universe](https://r-universe.dev/search) with an
  [overview](https://r-universe.dev/datasets) of
  [datasets](../help/data) included in R packages

The websites of these repositories include instructions how to install packages
from them, and repositories can be selected using 
[utils::setRepositories()](../help/setRepositories). The following code shows
how to install packages from [R-Forge](https://r-forge.r-project.org/) as an
example:

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
# Select packages that are currently not installed or not functional
pkgs_install <- pkgs_new[!vapply(X = pkgs_new, FUN = requireNamespace,
                                 FUN.VALUE = logical(1), quietly = TRUE)]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(),
                   repos = "http://R-Forge.R-project.org", dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

Mirror websites (mirrors) are websites hosted in various parts of the world with
the same content as the main website. Using a nearby mirror allows for faster
downloads. Mirrors of repositories can be selected using
[utils::setRepositories()](../help/setRepositories).

CRAN [mirrors](https://cran.r-project.org/mirrors.html), with information about
their [status](https://cran.r-project.org/mirmon_report.html) that is also
available from within R through [utils::getCRANmirrors()](../help/getCRANmirrors),
can be selected using [utils::chooseCRANmirror()](../help/chooseCRANmirror).
However, RStudio
[uses](https://docs.posit.co/ide/user/ide/guide/environments/r/packages.html#primary-repository)
the [RStudio CRAN mirror](https://cran.rstudio.com) with its own global
distribution, see [getOption("repos")](../help/getOption).

Bioconductor [mirrors](https://bioconductor.org/about/mirrors/), with information
about their [status](https://bioconductor.org/dashboard/#mirror_status), can be
selected using [BiocManager::repositories()](../help/repositories) or
[chooseBioCmirror()](../help/chooseBioCmirror). In RStudio, the RStudio CRAN
mirror is used (see above), which is signalled by the message
`'getOption("repos")' replaces Bioconductor standard repositories, see 'help("repositories", package = "BiocManager")' for details`.

### Updating packages
Updating out-of-date packages prevents compatibility issues between
already-installed and newly-installed packages. For R packages from CRAN,
versions can be compared using [diffify](https://diffify.com/R) and a
chronological overview of changes is available at
[CRANberries](https://dirk.eddelbuettel.com/cranberries/). To get the version
number of an installed package, run
[utils::packageVersion](../help/packageVersion)`("<pkgname>")`. 

The following code can be used to install the latest version of packages from
[CRAN](https://cran.r-project.org/web/packages/index.html) (but note this
**changes** the version of already-installed packages, which might be undesirable):

```{r, eval = FALSE}
utils::update.packages(lib.loc = .libPaths(), ask = TRUE, dependencies = NA,
                       verbose = getOption("verbose"), quiet = FALSE,
                       checkBuilt = TRUE, type = getOption("pkgType"))
```

The following code can be used to update packages to a specific BioConductor
release (here version 3.22), see the section [BioConductor] above:

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = character(), lib = .libPaths(), dependencies = NA,
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = TRUE, ask = TRUE, checkBuilt = TRUE, force = FALSE,
                     version = "3.22")
```

### Installing old versions
Installing old versions of a package might require installing Rtools to build
the packages from source, see the section 'Rtools' in the vignette
*Installing R, Rtools and RStudio*: `vignette("install_r", package = "checkrpkgs")`.

The following code can be used to install an old version of a package from CRAN,
using package [remotes](https://CRAN.R-project.org/package=remotes):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_version(package = <pkgname>, version = "1.23", dependencies = NA,
                         upgrade = "ask", quiet = FALSE, build_vignettes = TRUE,
                         lib = .libPaths(), verbose = getOption("verbose"))
```

Alternatively, visit the installation page of a package from CRAN, go to
`Downloads` > `Old sources` > `<pkgname> archive` and find the appropriate URL
pointing to an older version to install it using base R. For example, to install
version 1.40 of package [deSolve](https://CRAN.R-project.org/package=deSolve):

```{r, eval = FALSE}
install.packages(pkgs = "https://cran.r-project.org/src/contrib/Archive/deSolve/deSolve_1.40.tar.gz",
                 lib = .libPaths(), repos = NULL, dependencies = NA,
                 type = getOption("pkgType"), verbose = getOption("verbose"),
                 quiet = FALSE)
```

Older versions of packages from Bioconductor can be installed by changing the
value of argument `version` of `BiocManager::install()`, see the
[overview](https://bioconductor.org/about/release-announcements/) of
Bioconductor versions with the corresponding R version. Information on these
older packages can be found by visiting the appropriate version of Bioconductor,
e.g., `https://bioconductor.org/packages/3.21/BiocViews.html`.

### Troubleshooting
- If a package appears not to be installed when you want to use a function from
  it (e.g., you get the error `could not find function "<funcname>"`), remember
  you need to run [library](../help/library)`(<pkgname>)` to be able to use its
  functions.
  
- If a package is not functional, re-install it using argument `force = TRUE`.
  You can also use
  [checkrpkgs::get_details_pkgs](../help/get_details_pkgs)`(pkgs = <pkgname>)`
  to check which dependencies and other system requirements it has and then use
  [checkrpkgs::check_pkgs](../help/check_pkgs)`(pkgs = <pkgnames>)` to check if
  all the dependencies are installed and functional.

- If the warnings `lib = <pkgname> is not writeable` or
  `'lib' element <element from .libPaths()> is not a writable directory` occur,
  you probably forgot to run R (or RStudio) as administrator. Close R (RStudio),
  right-click on the R or RStudio icon and select `Run as administrator` to
  start it as administrator.

- If the question
  `Do you want to install from sources the packages which need compilation?` is
  asked (in a new window), accompanied by the remark
  `There are binary versions available but the source versions are later` and an
  overview in the console of the binary and source versions indicating if these
  need compilation, you should choose `No` if you have *not* installed
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/): you will then get
  slightly less up to date package versions but a faster installation.
  
  If you want the latest version without using Rtools, you can try again a few
  days later: it usually takes a bit longer for the binaries (i.e., the
  package versions that do not have to be build from source) from CRAN and
  Bioconductor to be updated than for the source versions used by Rtools.
  
  If you have installed Rtools you can choose `Yes` to install the latest
  package versions by building them from source. To install Rtools, see the
  section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
  `vignette("install_r", package = "checkrpkgs")`.

- If the warning `package <pkgname> was built under R version 'x.y.z'` occurs,
  you installed a binary package (i.e., not by building from source) that was
  prepared ('compiled') for an earlier version of R than the version of R you
  are currently using (run [getRversion()](../help/R.Version) to see your
  current R version). The warning is issued because packages are not tested on
  versions of R that are older than the version they were built on. Therefore it
  is best to update R when installing packages.

- The warning `package '<pkgname>' is not available (for R version x.y.z)` can
  have many reasons. First double-check the package name (which is
  case-sensitive). Then check possible other reasons mentioned in this
  [stackoverflow answer](https://stackoverflow.com/a/25721890/32365738).

- If errors occur when loading packages that require Java, make sure the 64-bit
  [version of Java](https://www.java.com/download/manual.jsp) is installed on
  64-bit PCs.


## Information about packages
Information about packages can be obtained from the internet before the packages
are installed, as well as from within R the packages are installed.

### Not-yet-installed packages
The [reference manual](https://cran.r-project.org/manuals.html) is a PDF-file
with all the help files of the standard and recommended packages. The manuals
and help pages of all packages from CRAN can be searched
[online](https://search.r-project.org/), or from within R using
[utils::RSiteSearch()](../help/RSiteSearch). In addition,
[utils::help.search()](../help/help.search) can be used to search the help
system using fuzzy matching or regular expressions, which can be disabled by
setting argument `agrep` to `FALSE` to search faster and return fewer results:
`utils::help.search(..., agrep = FALSE)`.

[tools::CRAN_package_db()](../help/CRAN_package_db) gives information about
packages available from [CRAN](https://cran.r-project.org/web/packages/index.html);
`BiocManager::available()` gives the names of packages available from
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html).

The source code of base R packages can be obtained from
[GitHub](https://github.com/r-devel/r-svn/) (see the section [Repositories]
below), and installation pages of packages on CRAN frequently contain links to
GitHub pages where their source code can be viewed.

#### Package dependencies
For (not necessarily installed) packages from CRAN, use
[tools::package_dependencies](../help/package_dependencies)`(packages = "<pkgname>", recursive = TRUE)`
to see dependencies (i.e., which packages are required by package `<pkgname>`) and
[tools::dependsOnPkgs](../help/dependsOnPkgs)`(pkgs = "<pkgname>", recursive = TRUE)`
to see reverse dependencies (i.e., which packages require package `<pkgname>`).
`NULL` is returned for packages that are not found, wheres `character(0)` is
returned for packages that do not have any dependencies.

### Already-installed packages
The locations where packages have been installed can be obtained with
[.libPaths()](../help/.libPaths), and the names of all installed packages can be
obtained with
[list.files](../help/list.files)`(path = .libPaths(), recursive = FALSE)`. The
repository from which a package was installed can be obtained from the
`Repository` and `URL` fields of package descriptions:
[checkrpkgs::get_details_pkgs](../help/get_details_pkgs)`(pkgs = <pkgname>)`.
The output of `checkrpkgs::get_details_pkgs()` can be used to re-install
packages after installing a new version of R.

Information about a package and its functions is available from within R after
the package has been installed and loaded (i.e.,
[library](../help/library)`(<pkgname>)` has been run):

- Citation: [utils::citation](../help/citation)`("<pkgname>")`, with
  `utils::citation()` to cite R itself.
- Function overview: `ls(getNamespace("<pkgname>"))` returns a character vector
  with the function names (ignoring names that start with a dot, which by
  convention are for internal use in packages;
  `ls(getNamespace("<pkgname>"), all.names = TRUE)` includes those function
  names in the returned character vector);
  `help(package = "<pkgname>")` gives an overview with links to their help-pages
  if there is a file `<pkgname>.R` in folder `<pgkname>\R`.
- Help page: `help(topic = "<pkgname>")`.
- Version: [utils::packageVersion](../help/packageVersion)`("<pkgname>")`.
- Vignettes: show them in a browser through
  [utils::browseVignettes](../help/browseVignettes)`(package = "<pkgname>")`
  or list them with [utils::vignette](../help/vignette)`(package = "<pkgname>")`.

Information about functions, methods, and classes can also be obtained (see also
the section [Getting the source code] below):

- Arguments: `args("<funcname>")`.
- Help page: `help("<funcname>")`;  
  indicate the package to distinguish functions with the same name from
  different packages: `help("<funcname>", package = "<pkgname>")`;  
  use quotes around the name of an operator to get its help page: `help("%in%")`.
- Methods: for a generic class:
  [utils::methods](../help/methods)`(class = "<classname>")`;  
  for a generic function: [utils::methods](../help/methods)`("<funcname>")`;  
  for S3-methods: `attr(utils::methods(class = "<classname>"), "info")`;  
  for S4-methods:
  `methods::showMethods(classes = "<classname>", where = getNamespace("<pkgname>"))`.
- Objects (including functions) whose name contains a certain string:
  [utils::apropos](../help/apropos)`("<string>")`.


### Which packages are used?
The function [loadedNamespaces()](../help/loadedNamespaces) shows which packages
are loaded. To see which packages are used in a script, look for `::`, `:::`,
[library](../help/library), [require](../help/require), and `namespace` (e.g.,
[loadNamespace](../help/loadNamespace),
[requireNamespace()](../help/requireNamespace)).

To see which packages are mentioned in comments, also look for:

- `conductor`: [Bioconductor](https://bioconductor.org/),
  [Neuroconductor](https://neuroconductor.org/)
- `CRAN`: [CRAN](https://cran.r-project.org/)
- `forge`: [R-Forge](https://r-forge.r-project.org/)
- `Git`: [GitHub](https://github.com/)
- `package` and `libraries`
- `rOpenSci`: [rOpenSci](https://ropensci.org/packages/)
- `verse`: [R-multiverse](https://r-multiverse.org/),
  [R-universe](https://r-universe.dev/packages),
  [tidyverse](https://tidyverse.org/packages/), etc

## Getting the source code

### Acknowledgements
Partly based on:

- Bryan J. 2015. [Accessing R Source](https://github.com/jennybc/access-r-source).
- Ligges U. 2006. Accessing the sources.
  [RNews 6(4):43-45](https://journal.r-project.org/articles/RN-2006-035/).
- A [community answer](https://stackoverflow.com/a/19226817) from
  [StackOverflow](https://stackoverflow.com/).
- The help page for [utils::methods()](../help/methods)

### Repositories
The source code of the base R packages, for the development, current, and older
versions of R, is available at [CRAN](https://cran.r-project.org/src/base/)
and at the [SVN-project](https://svn.r-project.org/R/branches/). Searching the
source code of the development version is easiest using the
[GitHub mirror](https://github.com/r-devel/r-svn/tree/main/src/library) of
the SVN-project, also the documentation on
[searching GitHub](https://docs.github.com/en/search-github).

The source code of add-on packages from
[CRAN](https://cran.r-project.org/web/packages/index.html) can be searched at
[METACRAN](https://github.com/cran), an *unofficial* CRAN mirror. Alternatively,
download the source file from section `Downloads` of the CRAN page. The source
files have been compressed into a `tar` file so you have to extract the files
(right-click on the downloaded file and choose `extract all`).

The source code of add-on packages from [BioConductor](https://bioconductor.org/)
in the current version or in the development-version is available
[here](https://code.bioconductor.org/search/).

The source code of add-on packages from [GitHub](https://github.com/) can be
viewed directly on GitHub, or after downloading the code to your computer by
clicking the green `Code` button on the repository page (e.g.,
`https://github.com/JesseAlderliesten/checkrpkgs`), choosing `Download ZIP`, and
unzipping the downloaded file (right-click on the file and choose `extract all`).

### Basic method
The simplest way to obtain the source code of a function is to type the name of
the function, *without* the brackets, and press `Enter`. For example, to see
what happens when using [sd](../help/sd) to calculate the standard deviation:

```{r}
sd
```

Some special cases:

- To distinguish functions with the same name from different packages, specify
  the package followed by two colons: `<pkgname>::<funcname>`.
- For non-exported (i.e., internal) functions, specify the package followed by
  three colons: `<pkgname>:::<funcname>`. Non-exported functions should *not* be
  used in code because they might change.
- For operators such as [%in%](../help/match) which start with a symbol, use
  backticks (`) around the name:
  
```{r}
`%in%`
```

### getAnywhere
A more robust alternative to the [basic method][Basic method] outlined above is
to use [getAnywhere](../help/getAnywhere)`("<funcname>")`, which looks in more
places and finds internal functions without the need to specify in which package
a function is defined. Although the quotes around the function name are not
necessary when looking for the source code of normal functions, they are
required when looking for the source code of operators such as
[%in%](../help/match), such that it is most robust to always use them.

#### UseMethod
If `getAnywhere("<funcname>")` returns `UseMethod("<funcname>")`, the function
has different methods for different object classes and is
[S3-generic](https://cran.r-project.org/doc/manuals/R-intro.html#Object-orientation).
First use [methods](../help/methods)`("<funcname>")` to get an overview of the
available methods; then obtain the source code of a particular method by using
running the code `getAnywhere("<function.class>")`, where `<function.class>` is
an item from the overview that was returned by `methods("<funcname>")`. Using
`getAnywhere("<function.class>")` instead of simply running `"<function.class>"`
ensures this also works for functions that are not exported, which is indicated
in the overview of `methods(<funcname>)` by an asterisk and the remark
`Non-visible functions are asterisked`.

For example, `UseMethod("mean")` in the output of `getAnywhere("mean")`
indicates `mean` is an S3-generic:
```{r, eval = TRUE}
getAnywhere("mean")
```

First use `methods("mean")` to get an overview of the available methods:
```{r, eval = TRUE}
methods("mean")
```

The output shows, among others, the methods `mean.Date` and `mean.default`.
Then use `getAnywhere("mean.Date")` to see the source code of the method `mean`
used with objects of class `Date`. 

```{r, eval = TRUE}
getAnywhere("mean.Date")
```

The source code of the method `mean` used with objects of classes not listed in
the output of `methods("mean")` is given by `getAnywhere("mean.default")`:

```{r, eval = TRUE}
getAnywhere("mean.default")
```

#### standardGeneric
If `getAnywhere("<funcname>")` returns `standardGeneric("<funcname>")`, the
function has different S4-methods (see
`help(topic = "Introduction", package = "methods")`) for different
object classes. Use [showMethods](../help/showMethods)`("<funcname>")` to get an
overview of the available methods in all *loaded* packages, or use
`<pkgname>:::<funcname>` to get an overview of the available methods from package
`<pkgname>`. Then provide the function name as argument `f` and the *entire line*
of the selected method as argument [signature](../help/signature) to function
[getMethod()](../help/getMethod), i.e.,
`getMethod(f = "<funcname>", signature = c(target = "<classname>", current = "<classname>"))`
to obtain the source code of a particular method.

For example, if the Matrix package is installed and loaded (i.e.,
`library(Matrix)` has been run), `getAnywhere("all.equal")` returns matches for
package Matrix and for package base. Selecting the match from package Matrix
(i.e., `getAnywhere("all.equal")[1]`) returns, among others,
`standardGeneric("all.equal")` indicating that `all.equal()` is an S4 function.
`showMethods("all.equal")` gives an overview of the different methods. Then
providing the function name as argument `f` and the *entire line* of the
selected method as argument `signature` to function `getMethod()`, i.e.,
``,
returns the source code for that particular method:

```{r, eval = TRUE}
library(Matrix)
getMethod(f = "all.equal", signature = c(target = "vector", current = "Matrix"))
```

#### .Internal or .Primitive
If `getAnywhere("<funcname>")` returns `.Internal` or `.Primitive`, the function
is internal or primitive. The source code of such functions can be viewed at
code [repositories][Repositories], or on your computer if you have installed R
[from source](https://cran.r-project.org/sources.html) using Rtools (see the
section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
`vignette("install_r", package = "checkrpkgs")`).

Locate the file `src/main/names.c` and look in the first column for the name of
the R function to find the appropriate c-entry which is given in the second
column of that file. Then either search for that c-entry using the GitHub
search, or manually locate the c-file (the name of the file is the c-entry
without the prefix `do_`) in `src/main` to get the file with the source code.

For example, `getAnywhere("matrix")` shows, among others, the line
`.Internal(matrix(data, nrow, ncol, byrow, dimnames, missing(nrow), missing(ncol)))`.
The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `matrix` in the first column:
`{"matrix",	do_matrix,	0,	11,	7,	{PP_FUNCALL, PREC_FN,	0}}`
Searching `src/main` for `matrix` gives file
[array.c](https://github.com/r-devel/r-svn/blob/main/src/main/array.c) as one of
the results. That file contains the source code of `matrix`. Similarly,
`getAnywhere("log10")` shows `function (x)  .Primitive("log10")`. The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `log10` in the first column:
`{"log10",	do_log1arg,	10,	1,	1,	{PP_FUNCALL, PREC_FN,	0}}`. Searching
`src/main` for `log1arg` gives file
[arithmetic.c](https://github.com/r-devel/r-svn/blob/main/src/main/arithmetic.c)
as one of its results. That file contains the source code of `log10`.

Sometimes an R-function is defined in a file that defines multiple functions and
thus has a general name. Then the file with c-code will have a similar general
name. For example, the source code of `make.names()` is defined in
`src/library/base/R/character.R` which contains (among others) the code
`.Internal(make.names(names, allow_))`, and the c-code for `do_makenames()` is
in file `src/main/character.c`.

#### .Call
If `getAnywhere("<funcname>")` returns code that contains `.Call(...)`, the
function contains a call to C or C++ code. Although the file `src/<pkgname>.h`
(e.g.,
[src/methods.h](https://github.com/r-devel/r-svn/blob/main/src/library/methods/src/methods.h)
if code is from package `methods`) contains an overview of the C or C++ code
included in a package, your best bet for finding the source code is
[searching](https://docs.github.com/en/search-github) the relevant part (e.g.,
`https://github.com/r-devel/r-svn/tree/main/src/library/methods/src` for code
from package `methods`) of the [GitHub](https://github.com/r-devel/r-svn/)
repository, or installing the package from source (see the section 'Rtools' in
the vignette *Installing R, Rtools and RStudio*:
`vignette("install_r", package = "checkrpkgs")`) and searching in the `src`
folder of the downloaded code.

## Documentation and help
- [Instructions](https://bioconductor.org/install/) on installing
  [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
  packages and [documentation](https://docs.r-universe.dev/bioconductor/) on the
  collaboration between BioConductor and the
  [R-universe](https://r-universe.dev/search)
- [Documentation](https://docs.r-universe.dev/) from the
  [R-universe](https://r-universe.dev/search)
- [Search engines](https://cran.r-project.org/search.html) specific for R
- Section
  [Add-on packages](https://cran.r-project.org/doc/manuals/R-admin.html#Add_002don-packages)
  in the
  [R Installation and Administration manual](https://cran.r-project.org/doc/manuals/r-release/R-admin.html)
  manual
- Section 'Documentation and help' in the vignette
  *Installing R, Rtools and RStudio*:  
  `vignette("install_r", package = "checkrpkgs")`
- Section [Troubleshooting] above
- The help-page [checkrpkgs::check_pkgs()](../help/check_pkgs)
