---
title: "Instructions about R packages"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Instructions about R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction and notation
This vignette contains information on installing and getting information about R
packages. It also explains how to obtain the source code of R functions.

In this vignette, `<pkgname>` is used to refer to a package name that should be
filled in to get working code, e.g., `<pkgname>` should be replaced with `grid`
if you want to obtain information about package `grid`, and with `stats` if you
want to obtain information about package `stats`. Similarly, `<funcname>` is
used to refer to a function name that should be filled in to get working code.


## R packages
An R
[package](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-is-the-difference-between-package-and-library_003f)
is a standardized collection of material extending R by providing code, data, or
documentation.

The
[base](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
packages are always installed together with R. The
[recommended](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-from-CRAN)
packages are installed with binary distributions of R. Together, the base and
recommended packages are the 'high priority' packages. To see which
high-priority packages are currently installed, run
`sort(unname(installed.packages(priority = "high")[, "Package"]))`. Since R
4.4.0, `tools::standard_package_names()` contains a list with the names of these
packages. Package 'translations' is not a recommended package, but will be
installed with R if that option is set during the installation.

### Installing packages
To install a package, run R (or RStudio) as administrator: right-click on the R
(RStudio) icon and select 'Run as administrator'. Packages can be obtained from
several websites (see below). After installing a package, you need to run
[library](../help/library)`(<pkgname>)` to be able to use the functions of that
package.

#### BioConductor
A [Bioconductor](https://bioconductor.org/) release uses specific versions of
packages from [CRAN](https://cran.r-project.org/web/packages/index.html) and
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html) that
are consistent with each other and with a
[specific version](http://bioconductor.org/about/release-announcements/#release-versions)
of R, preventing version conflicts. Twice a year a new BioConductor version is
released. Bioconductor also has thematic package collections known as
[BiocViews](https://bioconductor.org/packages/release/BiocViews.html).

The following code can be used to install packages from BioConductor release
[version](http://bioconductor.org/about/release-announcements/#release-versions)
3.22, using package [BiocManager](https://CRAN.R-project.org/package=BiocManager):

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = pkgs_new, lib = .libPaths(), dependencies = NA,
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = FALSE, ask = TRUE, checkBuilt = TRUE,
                     force = FALSE, version = "3.22")
```

#### CRAN
The [Comprehensive R Archive Network](https://cran.r-project.org/) (CRAN) is the
main repository of R [packages](https://cran.r-project.org/web/packages/index.html).
The following code can be used to install packages from CRAN:

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
pkgs_install <- pkgs_new[!vapply(X = pkgs_new, FUN = requireNamespace,
                                 FUN.VALUE = logical(1), quietly = TRUE)]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

CRAN has thematic package collections known as
[task views](https://cran.r-project.org/web/views/). To install all core packages
of a task view, install package [ctv](https://CRAN.R-project.org/package=ctv)
and run `ctv::install.views("<taskviewname>", coreOnly = TRUE)`. To update these
packages, use `ctv::update.views("<taskviewname>", coreOnly = TRUE)`.

Packages from CRAN that have been recently archived are available at
[CRANhaven](https://www.cranhaven.org/).

#### Github
The following code can be used to install packages from
[GitHub](https://github.com/), using package 
[remotes](https://CRAN.R-project.org/package=remotes):

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = NULL, dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_github(repo = pkgs_new, dependencies = NA, upgrade = FALSE,
                        force = FALSE, quiet = FALSE, build_vignettes = TRUE,
                        lib = NULL, verbose = getOption("verbose"))
```

#### Mirror websites
Mirror websites (mirrors) are websites hosted in various parts of the world with
the same content as the main website. Using a nearby mirror allows for faster
downloads. Mirrors, or repositories more generally, can be selected using
[utils::setRepositories()](../help/setRepositories). However, 
[RStudio](https://posit.co/products/open-source/rstudio/?sid=1)
[uses](https://docs.posit.co/ide/user/ide/guide/environments/r/packages.html#primary-repository)
the [RStudio CRAN mirror](https://cran.rstudio.com) with its own global
distribution, see [getOption("repos")](../help/getOption).

CRAN [mirrors](https://cran.r-project.org/mirrors.html), with information about
their [status](https://cran.r-project.org/mirmon_report.html) that is also
available from within R through [getCRANmirrors()](../help/getCRANmirrors), can
be selected using [chooseCRANmirror()](../help/chooseCRANmirror).

Bioconductor [mirrors](https://bioconductor.org/about/mirrors/), with information
about their [status](https://bioconductor.org/dashboard/#mirror_status), can be
selected using [BiocManager::repositories()](../help/repositories) or
[chooseBioCmirror()](../help/chooseBioCmirror). In RStudio, the RStudio CRAN
mirror is used (see above), which is signalled by the message
`'getOption("repos")' replaces Bioconductor standard repositories, see 'help("repositories", package = "BiocManager")' for details`.

#### Other repositories

- [Neuroconductor](https://neuroconductor.org/list-packages/all)
- [R-Forge](https://r-forge.r-project.org/) with a 
  [GitHub mirror](https://github.com/r-forge) and thematic package
  [collections](https://r-forge.r-project.org/softwaremap/trove_list.php) 
- [R-multiverse](https://r-multiverse.org/overview.html) with
  thematic [collections](https://r-multiverse.org/topics/)
- [rOpenSci](https://ropensci.org/packages/all/) with thematic package
  [collections](https://ropensci.org/packages/)
- [R universe](https://r-universe.dev/search) with an
  [overview](https://r-universe.dev/datasets) of
  [datasets](../help/data) included in R packages

### Installing old versions
Installing old versions of a package might require installing Rtools to build
the packages from source, see the section 'Rtools' in the vignette
*Installing R, Rtools and RStudio*: `vignette("install_r", package = "checkrpkgs")`.

Older versions of packages from Bioconductor can be installed by changing the
value of argument `version` of `BiocManager::install()`, see the
[overview](https://bioconductor.org/about/release-announcements/) of
Bioconductor versions with the corresponding R version. Information on these
older versions can be found by visiting the appropriate version of Bioconductor,
e.g., `https://bioconductor.org/packages/3.21/BiocViews.html`.

The following code can be used to install an old version of a package from CRAN,
using package [remotes](https://CRAN.R-project.org/package=remotes):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_version(package = <pkgname>, version = "1.23", dependencies = NA,
                         upgrade = "ask", quiet = FALSE, build_vignettes = TRUE,
                         verbose = getOption("verbose"))
```

Alternatively, at the installation page of a package from CRAN go to
`Downloads` > `Old sources` > `<pkgname> archive` and find the appropriate URL
pointing to an older version to install it using base R. For example, to install
version 1.40 of package [deSolve](https://CRAN.R-project.org/package=deSolve):

```{r, eval = FALSE}
install.packages(pkgs = "https://cran.r-project.org/src/contrib/Archive/deSolve/deSolve_1.40.tar.gz",
                 lib = .libPaths(), repos = NULL, dependencies = NA,
                 type = "source", verbose = getOption("verbose"),
                 quiet = FALSE)
```

### Updating packages
Updating out-of-date packages prevents compatibility issues between
already-installed and newly-installed packages. To get the version number of an
installed package, run
[utils::packageVersion](../help/packageVersion)`("<pkgname>")`. For R packages
from CRAN, versions can be compared using [diffify](https://diffify.com/R) and a
chronological overview of changes is available at
[CRANberries](https://dirk.eddelbuettel.com/cranberries/).

The following code can be used to update packages to a specific
BioConductor release (here version 3.22), see the section [BioConductor] above:

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = NULL, dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = character(), lib = NULL, verbose = FALSE,
                     update = TRUE, ask = TRUE, checkBuilt = TRUE,force = FALSE,
                     version = "3.22")
```

Alternatively, the following code can be used to install the latest version of
packages from [CRAN](https://cran.r-project.org/web/packages/index.html) (but
note this *changes* the version of already-installed packages, which might be
undesirable):

```{r, eval = FALSE}
utils::update.packages(lib.loc = .libPaths(), ask = TRUE, dependencies = NA,
                       verbose = getOption("verbose"), quiet = FALSE,
                       checkBuilt = TRUE, type = getOption("pkgType"))
```

### Troubleshooting
- If a package appears not to be installed when you want to use a function from
  it, remember you need to run [library](../help/library)`(<pkgname>)` to be
  able to use a function from package `<pkgname>`.

- If a package is not functional, re-install it using argument `force = TRUE`.
  You can also use
  [checkrpkgs::get_details_pkgs](../help/get_details_pkgs)`(pkgs = <pkgname>)`
  to check which dependencies and other system requirements it has and then use
  [checkrpkgs::check_pkgs](../help/check_pkgs)`(pkgs = c(<pkgnames>))` to check
  if all the dependencies are installed and functional.

- If the warning `lib = <pkgname> is not writeable` occurs, you probably forgot
  to run R (or RStudio) as administrator. Close R (RStudio), right-click on the
  R or RStudio icon and select `Run as administrator` to start it as
  administrator.

- If the question
  `Do you want to install from sources the packages which need compilation?` is
  asked, possibly accompanied by the remark
  `There are binary versions available but the source versions are later`, you
  should choose `No` if you have *not* installed
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/): you will then get
  slightly less up to date package versions but a faster installation.
  
  If you want the latest version without using Rtools, you can try again a few
  days later: it usually takes a bit longer for the CRAN binaries (i.e., the
  package versions that do not have to be build from source) to be updated than
  for the source versions used by Rtools.
  
  If you have installed Rtools you can choose `Yes` to install the latest
  package versions by building them from source. To install Rtools, see the
  section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
  `vignette("install_r", package = "checkrpkgs")`.

- If the warning `package <pkgname> was built under R version 'x.y.z'` occurs,
  you installed a binary package (i.e., not by building from source) that was
  compiled for an earlier version of R than the version of R you are currently
  using (run [R.version.string](../help/R.Version) to see your current R version).
  The warning is issued because packages are not tested on versions of R that
  are older than the version they were built on. Therefore it is best to update
  R when installing packages.

- The warning `package '<pkgname>' is not available (for R version x.y.z)` can
  have many reasons. First double-check the package name (which is
  case-sensitive). Then check possible other reasons mentioned in this
  [stackoverflow answer](https://stackoverflow.com/a/25721890/32365738).

- If errors occur when loading packages that require Java, make sure the 64-bit
  [version of Java](https://www.java.com/download/manual.jsp) is installed on
  64-bit PCs.


## Information about packages

### Not-yet-installed packages
The [reference manual](https://cran.r-project.org/manuals.html) is a PDF-file
with all the help files of the standard and recommended packages. The manuals
and help pages of all packages from CRAN can be searched
[online](https://search.r-project.org/) or from within R using
[utils::RSiteSearch()](../help/RSiteSearch).
[utils::help.search()](../help/help.search) searches uses fuzzy matching or
regular expressions while searching the help system, which can be disabled by
setting argument `agrep` to `FALSE` to search faster and return fewer results:
`utils::help.search(..., agrep = FALSE)`.

[tools::CRAN_package_db()](../help/CRAN_package_db) gives information about
packages available from [CRAN](https://cran.r-project.org/web/packages/index.html);
`BiocManager::available()` gives the names of packages available from
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html).

The source code of base R packages can be obtained from
[GitHub](https://github.com/r-devel/r-svn/) (see the section
[Getting the source code] below), and the installation pages of packages on CRAN
frequently contain links to the GitHub pages where their source code can be viewed.

#### Package dependencies
Use
[tools::package_dependencies](../help/package_dependencies)`(packages = "<pkgname>", recursive = TRUE)`
to see dependencies and
[tools::dependsOnPkgs](../help/dependsOnPkgs)`(pkgs = "<pkgname>", recursive = TRUE)`
to see reverse dependencies.

### Already-installed packages
The locations where packages have been installed can be obtained with
[tools::.libPaths()](../help/.libPaths), and the names of all installed packages
can be obtained with
[list.files](../help/list.files)`(path = .libPaths(), recursive = FALSE)`. The
repository from which a package was installed can be obtained from the
`Repository` and `URL` fields of package descriptions:
[checkrpkgs::get_details_pkgs](../help/get_details_pkgs)`(pkgs = <pkgname>)`.
The output of `checkrpkgs::get_details_pkgs()` can be used to re-install
packages after installing a new version of R.

Information about a package and its functions is available from within R after
the package has been installed and loaded (i.e.,
[library](../help/library)`(<pkgname>)` has been run):

- Citation: [utils::citation("<pkgname>")](../help/citation), with
  `utils::citation()` to cite R itself.
- Function overview: `ls(getNamespace("<pkgname>"))` gives their names as a
  character vector;  
  `help(package = "<pkgname>")` gives an overview with links to their help-pages
  (does not work if there is no file `<pkgname>.R` in folder `<pgkname>\R`).
- Help page: `help(topic = "<pkgname>")`.
- Version: [utils::packageVersion](../help/packageVersion)`("<pkgname>")`.
- Vignettes: show them in a browser:
  [browseVignettes](../help/browseVignettes)`(package = "<pkgname>")`; or list
  them with [vignette](../help/vignette)`(package = "<pkgname>")`.

Information about functions, methods, and classes can also be obtained (see also
the section [Getting the source code] below):

- Arguments: `args("<funcname>")`.
- Help page: `help("<funcname>")`;  
  indicate the package to distinguish functions with the same name from
  different packages: `help("<funcname>", package = "<pkgname>")`;  
  use quotes around the name of an operator to get its help page: `help("%in%")`.
- Methods: for a generic class:
  [utils::methods](../help/methods)`(class = "<classname>")`;  
  for a generic function: [utils::methods](../help/methods)`("<funcname>")`;  
  for S3-methods: `attr(utils::methods(class = "<classname>"), "info")`;  
  for S4-methods:
  `methods::showMethods(classes = "<classname>", where = getNamespace("<pkgname>"))`.
- Objects (including functions) whose name contains a certain string:
  [utils::apropos](../help/apropos)`("<funcname>")`.


### Which packages are used?
The function [loadedNamespaces()](../help/loadedNamespaces) can be used to see
which packages are loaded. To see which packages are used in a script, look for
[library](../help/library), [require](../help/require), `namespace` (e.g.,
[loadNamespace](../help/loadNamespace),
[requireNamespace()](../help/requireNamespace)), `::`, and `:::`.

To see which packages are mentioned in comments, also look for:

- `conductor`: [Bioconductor](https://bioconductor.org/),
  [Neuroconductor](https://neuroconductor.org/)
- `CRAN`: [CRAN](https://cran.r-project.org/)
- `forge`: [R-Forge](https://r-forge.r-project.org/)
- `Git`: [GitHub](https://github.com/)
- `package`
- `rOpenSci`: [rOpenSci](https://ropensci.org/packages/)
- `verse`: [R-multiverse](https://r-multiverse.org/),
  [R-universe](https://r-universe.dev/packages),
  [tidyverse](https://tidyverse.org/packages/), etc

## Getting the source code

### Acknowledgements
Partly based on:

- Bryan J. 2015. [Accessing R Source](https://github.com/jennybc/access-r-source).
- Ligges U. 2006. Accessing the sources.
[RNews 6(4):43-45](https://journal.r-project.org/articles/RN-2006-035/).
- A post from [StackOverflow](https://stackoverflow.com/a/19226817).

### Repositories
The source code of the base R packages, for the current as well as earlier
versions of R, is available at [CRAN](https://cran.r-project.org/src/base/)
and at the [SVN-project](https://svn.r-project.org/R/). Searching the source
code is easiest using the [GitHub mirror](https://github.com/r-devel/r-svn/) of
the latter.

The source code of [BioConductor](https://bioconductor.org/) packages is
available [here](https://code.bioconductor.org/search/). The source of the
add-on packages from [CRAN](https://cran.r-project.org/web/packages/index.html)
is available at [METACRAN](https://github.com/cran), an *unofficial* CRAN mirror.

### Basic method
The simplest way to obtain the source code of a function is typing the name of
the function, *without* the brackets. E.g., [sd](../help/sd) returns
`function (x, na.rm = FALSE) sqrt(var(if (is.vector(x) || is.factor(x)) x else as.double(x), na.rm = na.rm))`.

Some special cases:

- For operators such as [%in%](../help/match) which start with a symbol, use
  backticks around the name.
- To distinguish functions with the same name from different packages, specify
  the package followed by two colons: `<pkgname>::<funcname>`.
- For non-exported (i.e., internal) functions, specify the package followed by
  three colons: `<pkgname>:::<funcname>`. Non-exported functions should *not* be
  used directly: they might change.

### getAnywhere
A more robust alternative to the [basic method][Basic method] outlined above is
to use [getAnywhere](../help/getAnywhere)`("<funcname>")`, which looks in more
places and finds internal functions without the need to specify in which package
a function is defined. Although the quotes around the function name are not
necessary when looking for normal functions, they are required when looking for
the source code of operators such as [%in%](../help/match), such that it is most
robust to always use them.

#### UseMethod
If `getAnywhere("<funcname>")` returns `UseMethod("<funcname>")`, the function
has different methods for different object classes and is
[S3-generic](https://cran.r-project.org/doc/manuals/R-intro.html#Object-orientation).
Use [methods](../help/methods)`("<funcname>")` to get an overview of the
available methods and then `getAnywhere("<function.class>")`, where
`<function.class>` is an item from the overview returned by `methods("<funcname>")`,
to obtain the source code of a particular method.

For example, `getAnywhere("mean")` returns `function (x, ...) UseMethod("mean")`,
indicating it is an S3-generic. Using `methods("mean")` returns `mean.Date`,
`mean.default`, and other methods. `getAnywhere("mean.Date")` returns the source
code for the method used with objects of class 'Date', and
`getAnywhere("mean.default")` returns the source code for the default method,
i.e., the method used with objects of classes not listed in the output of
`methods("mean")`:

```{r, eval = FALSE}
function (x, trim = 0, na.rm = FALSE, ...) {
    if (!is.numeric(x) && !is.complex(x) && !is.logical(x)) {
        warning("argument is not numeric or logical: returning NA")
        return(NA_real_)
    }
    if (isTRUE(na.rm))
      x <- x[!is.na(x)]
    < omitted further output >
}
```

If the overview of `methods(<funcname>)` includes the remark
`Non-visible functions are asterisked`, it means that these functions are not
exported and you need to use `getAnywhere("<method.class>")` or
`<pkgname>:::<method.class>`.

#### standardGeneric
If `getAnywhere("<funcname>")` returns `standardGeneric("<funcname>")`, the
function has different S4-methods (see `methods::Introduction`) for different
object classes. Use [showMethods](../help/showMethods)`("<funcname>")` to get an
overview of the available methods *in all loaded packages*, or use
`<pkgname>:::<funcname>` to get an overview of the available methods from package
`<pkgname>`. Then provide the function name as argument `f` and the *entire line*
of the selected method as argument [signature](../help/signature) to function
[getMethod()](../help/getMethod), i.e.,
`getMethod(f = "<funcname>", signature = c(target = "<classname>", current = "<classname>"))`
to obtain the source code of a particular method.

For example, if the Matrix package is installed and loaded (i.e.,
`library(Matrix)` has been run), `getAnywhere("all.equal")` returns matches for
package Matrix and for package base. Selecting the match from package Matrix
(i.e., `getAnywhere("all.equal")[1]`) returns, among others,
`standardGeneric("all.equal")` indicating that 'all.equal' is an S4 function.
`showMethods("all.equal")` gives an overview of the different methods. Then
providing the function name as argument `f` and the *entire line* of the
selected method as argument `signature` to function `getMethod()`, i.e.,
`getMethod(f = "all.equal", signature = c(target = "vector", current = "Matrix"))`,
returns the source code for that particular method:

```{r, eval = FALSE}
function (target, current, ...) {
    .local <- function (target, current, check.type = check.class, 
        check.class = TRUE, check.attributes = TRUE, check.factors = FALSE, 
        ...) {
        msg <- .M.attr.all.equal(target, current, check.type = check.type, 
            check.class = check.class, check.attributes = check.attributes, 
            check.factors = check.factors, ...)
        < omitted further output >
    }
}
```

#### .Internal or .Primitive
If `getAnywhere("<funcname>")` returns `.Internal` or `.Primitive`, the function
is internal or primitive. The source code of such functions can be viewed at the
code repositories such as [GitHub](https://github.com/r-devel/r-svn/), or on
your computer if you have installed R
[from source](https://cran.r-project.org/sources.html) using Rtools (see the
section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
`vignette("install_r", package = "checkrpkgs")`).

Locate the file `src/main/names.c` and look in the first column for the name of
the R function to find the appropriate c-entry which is given in the second
column of that file. Then either search for that c-entry using the GitHub
search, or manually locate the c-file (the name of the file is the c-entry
without the prefix `do_`) in `src/main` to get the file with the source code.

For example, `getAnywhere("matrix")` shows, among others, the line
`.Internal(matrix(data, nrow, ncol, byrow, dimnames, missing(nrow), missing(ncol)))`.
The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `matrix` in the first column:
`{"matrix",	do_matrix,	0,	11,	7,	{PP_FUNCALL, PREC_FN,	0}}`
Searching `src/main` for `matrix` gives file
[array.c](https://github.com/r-devel/r-svn/blob/main/src/main/array.c) as one of
the results. That file contains the source code of `matrix`. Similarly,
`getAnywhere("log10")` shows `function (x)  .Primitive("log10")`. The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `log10` in the first column:
`{"log10",	do_log1arg,	10,	1,	1,	{PP_FUNCALL, PREC_FN,	0}}`. Searching
`src/main` for `log1arg` gives file
[arithmetic.c](https://github.com/r-devel/r-svn/blob/main/src/main/arithmetic.c)
as one of its results. That file contains the source code of `log10`.

#### .Call
If `getAnywhere("<funcname>")` returns code that contains `.Call(...)`, the
function contains a call to C or C++ code. Although the file `src/<pkgname>.h`
(e.g.,
[src/stats.h](https://github.com/r-devel/r-svn/blob/main/src/library/stats/src/stats.h)
if code is from package `stats`) contains an overview of the C or C++ code
included in a package, your best bet for finding the source code is using the
[GitHub](https://github.com/r-devel/r-svn/) search, or installing the package
from source and searching in the `src` folder of the downloaded code.


## Documentation and help
- [Instructions](https://bioconductor.org/install/) on installing
  [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
  packages
- [Search engines](https://cran.r-project.org/search.html) specific for R
- Section
  [Add-on packages](https://cran.r-project.org/doc/manuals/R-admin.html#Add_002don-packages)
  in the
  [R Installation and Administration manual](https://cran.r-project.org/doc/manuals/r-release/R-admin.html)
  manual
- Section 'Documentation and help' in the vignette
  *Installing R, Rtools and RStudio*:  
  `vignette("install_r", package = "checkrpkgs")`
- Section [Troubleshooting] above
- The help-page [checkrpkgs::check_pkgs()](../help/check_pkgs)
