---
title: "R packages"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction and notation
This vignette explains how to install R packages, how to get information about
to-be-installed and already-installed R packages, and how to get the source code
of R functions.

Text between angled brackets (`<...>`) is used to refer to
text that should be replaced with specific text to get working code or working
file paths. For example, `<pkg>` is used as a place holder to refer to a
package name and should be replaced with `utils` if you
want to get information about package `utils`, and with `methods` if you want to
get information about package `methods`. Similarly, `<func>` is used as a place
holder to refer to a function name that should be filled in with a specific
function name to get working code.

In this vignette, calls to functions are frequently written in the form
`<pkg>::<func>()`, to make clear which package is
used and, through the brackets, that a function is indicated. In normal code,
one would use `library(<pkg>)` followed by `<func>()`. For example, in
this vignette the notation `utils::citation()` is used to show how to cite R,
indicating that the function `citation()` is defined in
package `utils`. In normal code, one would use `library(utils)` followed by
`citation()` where that is needed.

## R packages
An R
[package](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-is-the-difference-between-package-and-library_003f)
is a standardized collection of material extending R by providing code, data, or
documentation.

The
[base](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
packages are always installed together with R. The
[recommended](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-from-CRAN)
packages are installed with binary distributions of R. Together, the base and
recommended packages are the 'high priority' packages. Since R 4.4.0,
`tools::standard_package_names()` contains a list with the names of these
packages. To see which high-priority packages are currently installed, run
`sort(unname(installed.packages(priority = "high")[, "Package"]))`. Package
'translations' is not a recommended package, but will be installed if that
option is set during the installation of R.

### Installing packages
To install a package, run R or RStudio as administrator: right-click on the R or
RStudio icon and select `Run as administrator`. Packages can be obtained from
several websites, called 'repositories', such as `CRAN`, `BioConductor`, and
`GitHub`, discussed in the next sections. After installing a package, you
need to [attach][attaching packages] the package to be able to use its
functions: run `library(<pkg>)`.

#### CRAN
The [Comprehensive R Archive Network](https://cran.r-project.org/) (CRAN) is the
main repository of R [packages](https://cran.r-project.org/web/packages/index.html).
The following code can be used to install packages from CRAN:

```{r, eval = FALSE}
pkgs_new <- c(<pkg>, <pkg>)
# Select packages from 'pkgs_new' that are not installed or not functional
pkgs_install <- pkgs_new[!vapply(X = pkgs_new, FUN = requireNamespace,
                                 FUN.VALUE = logical(1), quietly = TRUE)]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

CRAN has thematic package collections known as
[task views](https://cran.r-project.org/web/views/). To install all core packages
of a task view, install package [ctv](https://CRAN.R-project.org/package=ctv)
and run `ctv::install.views("<taskview>", coreOnly = TRUE)`. To update these
packages, use `ctv::update.views("<taskview>", coreOnly = TRUE)`.

`tools::CRAN_package_db()` gives information about packages available from
[CRAN](https://cran.r-project.org/web/packages/index.html), including the
`Description` and `Maintainer` fields not returned by
`utils::available.packages()`. `tools::CRAN_check_results()` gives information
about the current check status of CRAN packages. Packages from CRAN that have
been recently archived (for example because check issues were not adressed in
time) are available at [CRANhaven](https://www.cranhaven.org/).

#### BioConductor
The [Bioconductor](https://bioconductor.org/) repository has a new release every
six months. Each release contains specific versions of packages from
[CRAN](https://cran.r-project.org/web/packages/index.html)
and [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
that are consistent with each other and with a
[specific version](https://bioconductor.org/about/release-announcements/)
of R, preventing version conflicts between R packages.

The following code can be used to install packages from BioConductor release
[version](https://bioconductor.org/about/release-announcements/)
3.23. This code installs the
[BiocManager](https://CRAN.R-project.org/package=BiocManager) package from
[CRAN](https://cran.r-project.org/) that is then used to install packages from
`Bioconductor` and `CRAN` and, through `remotes::install_github()` (see the next
[section][GitHub]), from `GitHub`:

```{r, eval = FALSE}
pkgs_new <- c(<pkg>, <pkg>)
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = pkgs_new, lib = .libPaths(), dependencies = NA,
                     build_vignettes = TRUE,
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = FALSE, ask = TRUE, checkBuilt = TRUE,
                     force = FALSE, version = "3.23")
```

Bioconductor also has thematic package collections known as
[BiocViews](https://bioconductor.org/packages/release/BiocViews.html).

`utils::available.packages(fields = NULL, repos = BiocManager::repositories())`
gives information about packages available from
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html), and
`BiocManager::available()` gives their names.

#### Github
The following code can be used to install packages from
[GitHub](https://github.com/): it installs the
[remotes](https://CRAN.R-project.org/package=remotes) package that is needed to
install packages from GitHub, selects those elements of `pkgs_new` that give the
author name and repository name (e.g., `"JesseAlderliesten/checkrpkgs"`) or the
full URL to a package (e.g., `"https://github.com/JesseAlderliesten/checkrpkgs"`)
as required by `remotes::install_github()`, and installs those packages. To
match such names to package names as returned by `utils::installed.packages()`,
use `basename(pkgs_new)` to select the last part of those names:
`pkgs_new[!(basename(pkgs_new) %in% installed.packages()[, "Package"])]`.

```{r, eval = FALSE}
pkgs_new <- "JesseAlderliesten/checkrpkgs"
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_github(repo = grep(pattern = "/", x = pkgs_new, value = TRUE),
                        dependencies = NA, upgrade = "ask", force = FALSE,
                        quiet = FALSE, build_vignettes = TRUE, lib = .libPaths(),
                        verbose = getOption("verbose"))
```

#### Other repositories
Examples of other repositories for R packages are:

- [Neuroconductor](https://neuroconductor.org/list-packages/all)
- [R-Forge](https://r-forge.r-project.org/) with a 
  [GitHub mirror](https://github.com/r-forge) and thematic package
  [collections](https://r-forge.r-project.org/softwaremap/trove_list.php) 
- [R-multiverse](https://r-multiverse.org/overview.html) with
  thematic package [collections](https://r-multiverse.org/topics/)
- [rOpenSci](https://ropensci.org/packages/all/) with thematic package
  [collections](https://ropensci.org/packages/)
- [R universe](https://r-universe.dev/search) with an
  [overview](https://r-universe.dev/datasets) of
  datasets (see `help("data", package = "utils")`) included in R packages

The websites of these repositories include instructions how to install packages
from them, and repositories can be selected using `utils::setRepositories()`.
The following code shows how to install packages from
[R-Forge](https://r-forge.r-project.org/) as an example:

```{r, eval = FALSE}
pkgs_new <- c(<pkg>, <pkg>)
# Select packages from 'pkgs_new' that are not installed or not functional
pkgs_install <- pkgs_new[!vapply(X = pkgs_new, FUN = requireNamespace,
                                 FUN.VALUE = logical(1), quietly = TRUE)]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(),
                   repos = "https://r-forge.r-project.org/", dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

#### Mirror websites
Mirror websites ('mirrors') are websites hosted in various parts of the world
with the same content as the main website. Using a nearby mirror allows for
faster downloads. Mirrors of repositories can be selected using
`utils::setRepositories()`.

CRAN [mirrors](https://cran.r-project.org/mirrors.html), with information about
their [status](https://cran.r-project.org/mirmon_report.html) that is also
available from within R through `utils::getCRANmirrors()`, can be selected using
`utils::chooseCRANmirror()`. However, RStudio
[uses](https://docs.posit.co/ide/user/ide/guide/environments/r/packages.html#primary-repository)
the [RStudio CRAN mirror](https://cran.rstudio.com) with its own global
distribution, which is signalled by the message
`'getOption("repos")' replaces Bioconductor standard repositories, see 'help("repositories", package = "BiocManager")' for details`.

Bioconductor [mirrors](https://bioconductor.org/about/mirrors/), with information
about their [status](https://bioconductor.org/dashboard/#mirror_status), can be
selected using `BiocManager::repositories()` or `utils::chooseBioCmirror()`,
although the RStudio CRAN mirror is used in RStudio (see the preceding paragraph).

### Attaching packages
After installing a package, you need to attach the package to be able to use its
functions: run `library(<pkg>)`. If attaching fails without clear reason,'
setting environment variable `_R_TRACE_LOADNAMESPACE_` to a numerical value
(e.g., `Sys.setenv("_R_TRACE_LOADNAMESPACE_" = 4)`) will generate additional
messages on progress for non-standard packages (see the section `Tracing` in
`help("requireNamespace")`).

`loadedNamespaces()` gives the names of packages that are currently loaded,
`utils::sessionInfo()` also gives their versions, and
`sessioninfo::session_info()` provides more details and has the option to show
information about their dependencies. `options("defaultPackages")` gives the
names of packages that are attached by default when R starts up.

### Updating packages
Updating out-of-date packages prevents compatibility issues between
already-installed and newly-installed packages.

#### CRAN
For R packages from CRAN, versions can be compared using
[diffify](https://diffify.com/R) and a chronological overview of changes is
available at [CRANberries](https://dirk.eddelbuettel.com/cranberries/). To get
the version number of an installed package, run
`utils::packageVersion("<pkg>")`. `old.packages()` indicates which packages can
be updated.

The following code can be used to install the latest version of packages from
[CRAN](https://cran.r-project.org/web/packages/index.html) (this
**changes** the version of already-installed packages, which might be
undesirable):

```{r, eval = FALSE}
utils::update.packages(lib.loc = .libPaths(), ask = TRUE, dependencies = NA,
                       verbose = getOption("verbose"), quiet = FALSE,
                       checkBuilt = TRUE, type = getOption("pkgType"))
```

#### Bioconductor
`BiocManager::valid()` indicates which packages can be updated and also checks
for too new packages, taking the currently used version of Bioconductor (see
`BiocManager::version()`) into account. The following code can be used to update
packages to a specific BioConductor release (here version 3.23):

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = character(), lib = .libPaths(), dependencies = NA,
                     build_vignettes = TRUE, 
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = TRUE, ask = TRUE, checkBuilt = TRUE, force = FALSE,
                     version = "3.23")
```

### Installing old versions
Installing old versions of a package might require installing Rtools to build
the packages from source, see the section 'Rtools' in the vignette
*Installing R, Rtools and RStudio*: `vignette("install_r", package = "checkrpkgs")`.

#### CRAN
The following code can be used to install an old version of a package from CRAN,
using package [remotes](https://CRAN.R-project.org/package=remotes):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_version(package = "deSolve", version = "1.40", dependencies = NA,
                         upgrade = "ask", quiet = FALSE, build_vignettes = TRUE,
                         lib = .libPaths(), verbose = getOption("verbose"))
```

Alternatively, visit the installation page of a package from CRAN, go to
`Downloads` > `Old sources` > `<pkg> archive` and find the appropriate URL
pointing to an older version to install it using base R. For example, to install
version 1.40 of package [deSolve](https://CRAN.R-project.org/package=deSolve):

```{r, eval = FALSE}
install.packages(
  pkgs = "https://cran.r-project.org/src/contrib/Archive/deSolve/deSolve_1.40.tar.gz",
  lib = .libPaths(), repos = NULL, dependencies = NA,
  type = getOption("pkgType"), verbose = getOption("verbose"),
  quiet = FALSE)
```

#### Bioconductor
Older versions of packages from Bioconductor can be installed by changing the
value of argument `version` of `BiocManager::install()` (e.g.,
`BiocManager::install(pkgs = "deSolve", version = "3.22")` to indicate the
version of `deSolve` included in BioConductor version 3.22) but that only works
when using the version of R for that specific version of Bioconductor, see the
[overview](https://bioconductor.org/about/release-announcements/) of
Bioconductor versions with the corresponding R version. Information on these
older packages can be found by visiting the appropriate version of Bioconductor,
e.g., `https://bioconductor.org/packages/3.22/BiocViews.html`.

### Troubleshooting

#### Installing packages
- If the warnings `lib = <pkg> is not writeable` or
  `'lib' element <element from .libPaths()> is not a writable directory` occur,
  you probably forgot to run R (or RStudio) as administrator. Close it,
  right-click on the R (or RStudio) icon and select `Run as administrator` to
  start it as administrator.

- If the question
  `Do you want to install from sources the packages which need compilation?` is
  asked (in a new window), accompanied by the remark
  `There are binary versions available but the source versions are later` and an
  overview in the console of the binary and source versions indicating if these
  need compilation, you should choose `No` if you have **not** installed
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/): you will then get
  slightly less up-to-date package versions but a faster installation.
  
  If you want the latest version without using Rtools, you can try again a few
  days later: it usually takes a bit longer for the binaries (i.e., the
  package versions that do not have to be build from source) from CRAN and
  Bioconductor to be updated than for the source versions used by Rtools.
  
  If you have installed Rtools you can choose `Yes` to install the latest
  package versions by building them from source. To install Rtools, see the
  section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
  `vignette("install_r", package = "checkrpkgs")`.

- The warning `package '<pkg>' is not available (for R version x.y.z)` can
  have many reasons. First double-check the package name (which is
  case-sensitive). Then check possible other reasons mentioned in this
  [stackoverflow answer](https://stackoverflow.com/a/25721890/32365738).

- If installing packages fails, try with arguments `force = TRUE` to re-install
  possibly broken dependencies and with argument `build_vignettes = FALSE` to
  not install vignettes.

#### Using packages
- If a package appears not to be installed when you want to use a function from
  it (e.g., you get the error `could not find function "<func>"`), remember
  you need to run `library(<pkg>)` to be able to use its functions.

- To check that a package is installed and functional, use `library(<pkg)` or
  `requireNamespace(<pkg>)`. These functions do not allow vectors as input, such
  that the following code has to be used to check multiple packages:
  
  ``` {r, eval = FALSE}
  suppressPackageStartupMessages(
    !vapply(X = c(<pkg>, <pkg>), FUN = requireNamespace, FUN.VALUE = logical(1),
            lib.loc = NULL, quietly = FALSE)
  )
  ```

- If a package is not functional, re-install it using argument `force = TRUE` to
  re-install possibly broken dependencies. You can also use
  `checkrpkgs::get_details_pkgs(pkgs = <pkg>)` to check which dependencies
  and system requirements it has.

- If the warning `package <pkg> was built under R version 'x.y.z'` occurs,
  you installed a binary package (i.e., not by building from source) that was
  prepared ('compiled') for an earlier version of R than the version of R you
  are currently using (run `getRversion()` to see your current R version). The
  warning is issued because packages are **not** tested on versions of R that are
  older than the version they were built on. Therefore it is best to update R
  when installing packages.

- If errors occur when [attaching packages] that require Java, make sure the
  64-bit [version of Java](https://www.java.com/download/manual.jsp) is
  installed on 64-bit PCs.

## Information about packages
Information about packages can be obtained from the internet before the packages
are installed, as well as from within R after the packages are installed.

### Not-yet-installed packages
The [reference manual](https://cran.r-project.org/manuals.html) is a PDF-file
with all the help files of the standard and recommended packages. The manuals
and help pages of all packages from CRAN can be searched
[online](https://cran.r-project.org/search.html), or from within R using
`utils::RSiteSearch()`. In addition, `utils::help.search()` can be used to
search the help system using fuzzy matching or regular expressions, which can be
disabled by setting argument `agrep` to `FALSE` to search faster and return
fewer results: `utils::help.search(..., agrep = FALSE)`.

`utils::available.packages()` and `tools::CRAN_package_db()` give information
about packages available from
[CRAN](https://cran.r-project.org/web/packages/index.html);
`utils::available.packages(fields = NULL, repos = BiocManager::repositories())`
gives information about packages available from
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html), and
`BiocManager::available()` gives their names.

The source code of base R packages can be obtained from
[GitHub](https://github.com/r-devel/r-svn/) (see the section [Repositories]
below), and installation pages of packages on CRAN frequently contain links to
GitHub pages where their source code can be viewed.

For (not necessarily installed) packages from CRAN, use
`tools::package_dependencies(packages = "<pkg>", recursive = TRUE)` to see
dependencies (i.e., which packages are required by package `<pkg>`) and
`tools::dependsOnPkgs(pkgs = "<pkg>", recursive = TRUE)` to see reverse
dependencies (i.e., which packages require package `<pkg>`). `NULL` is
returned for packages that are not found, whereas `character(0)` is returned for
packages that do not have any dependencies. To see dependencies of packages from
other repositories (e.g., [GitHub](https://github.com/)), use package 
[pkgdepends](https://r-lib.github.io/pkgdepends/):

``` {r, eval = FALSE}
library(pkgdepends)
prop <- pkgdepends::new_pkg_deps("<repos>/<pkg>")
prop$solve()
prop$get_solution()$data
prop$draw()
```

### Already-installed packages
`utils::installed.packages()` gives details on installed packages. Argument
`fields` can be used to specify additional fields to extract from the package
DESCRIPTION, for example
```{r fields, eval = FALSE}
`fields = c("Repository", "Additional_repositories",
"URL", "GithubRepo", "GithubUsername", "SystemRequirements")`.
```
The `Repository` and `URL` fields show the repository from which a package was
installed and are conveniently shown by `sessioninfo::session_info()`, which
provides also has the option to show only information about selected packages
and their dependencies.

The locations where R installs packages, and where it looks for installed
packages, can be obtained with `.libPaths()`. The location where a particular
package is installed can be obtained with
`find.package(package = "<pkg>", lib.loc = NULL, verbose = TRUE)`, using
`verbose = TRUE` to get a warning if a package is found more than once. The
names of all installed packages can be obtained with
`list.files(path = .libPaths())`.

Information about a package and its functions, methods, and classes is available
from within R after the package has been installed and attached (i.e.,
`library(<pkg>)` has been run):

- Citation: `utils::citation("<pkg>")`, with `utils::citation()` to cite R
  itself.
- Function arguments: `args("<func>")`.
- Function help page: `help("<func>")`;
  indicate the package to distinguish functions with the same name from
  different packages: `help("<func>", package = "<pkg>")`;
  use quotes around the name of an operator to get its help page: `help("%in%")`.
- Function methods: for a generic class: `utils::methods(class = "<class>")`;
  for a generic function: `utils::methods("<func>")`;
  for S3-methods: `attr(utils::methods(class = "<class>"), "info")`;
  for S4-methods:
  `methods::showMethods(classes = "<class>", where = getNamespace("<pkg>"))`.
- Function overview: `ls(getNamespace("<pkg>"), all.names = TRUE)` returns a
  character vector with the function names (the default `all.names = FALSE`
  ignores names that start with a dot, which by convention are for internal use
  in packages; `help(package = "<pkg>")` gives an overview with links to their
  help-pages if there is a file `<pkg>.R` in folder `<pkg>\R`.
- Functions and other objects whose name contains a certain string:
  `utils::apropos("<string>")`.
- Installation path (i.e., where is the package installed):
  `find.package(package = "<pkg>", lib.loc = .libPaths(), verbose = TRUE)`.
- Version: `utils::packageVersion("<pkg>")`.
- Vignettes: show them in a browser through
  `utils::browseVignettes(package = "<pkg>")` or list them with
  `utils::vignette(package = "<pkg>")`.

### Which packages are used?
The function `loadedNamespaces()` shows which packages are loaded.
`getAnywhere(<func>)` shows in which package a function is defined. To see which
packages are used in a script, look for `::`, `:::`, `library`, `require`, and
`namespace` (e.g., `loadNamespace()`, `requireNamespace()`). However, various
packages have their own way to create dependencies on packages, see the overview
at [pak::scan_deps()](https://pak.r-lib.org/reference/scan_deps.html).

To see which packages are mentioned in comments, also look for:

- `conductor`: [Bioconductor](https://bioconductor.org/),
  [Neuroconductor](https://neuroconductor.org/)
- `CRAN`: [CRAN](https://cran.r-project.org/)
- `forge`: [R-Forge](https://r-forge.r-project.org/)
- `Git`: [GitHub](https://github.com/)
- `package` and `libraries`
- `rOpenSci`: [rOpenSci](https://ropensci.org/packages/)
- `verse`: [R-multiverse](https://r-multiverse.org/),
  [R-universe](https://r-universe.dev/packages),
  [tidyverse](https://tidyverse.org/packages/), etc

## Getting the source code

### Acknowledgements
Partly based on:

- Bryan J. 2015. [Accessing R Source](https://github.com/jennybc/access-r-source).
- Ligges U. 2006. Accessing the sources.
  [RNews 6(4):43-45](https://journal.r-project.org/articles/RN-2006-035/).
- A [community answer](https://stackoverflow.com/a/19226817) from
  [StackOverflow](https://stackoverflow.com/).
- The help page for `utils::methods()`.

### Repositories
The source code of the base R packages is available at
[CRAN](https://cran.r-project.org/src/base/)
and at the [SVN-project](https://svn.r-project.org/R/branches/). Searching the
source code of the development version is easiest using the
[GitHub mirror](https://github.com/r-devel/r-svn/tree/main/src/library) of
the SVN-project, see also the documentation on
[searching GitHub](https://docs.github.com/en/search-github).

The source code of packages from
[CRAN](https://cran.r-project.org/web/packages/index.html) can be searched at
[METACRAN](https://github.com/cran), an *unofficial* CRAN mirror. Alternatively,
download the source file from section `Downloads` of the CRAN page. The source
files have been compressed into a `tar` file so you have to extract the files
(right-click on the downloaded file and choose `extract all`).

The source code of add-on packages from [BioConductor](https://bioconductor.org/)
in the current version or in the development-version is available
[here](https://code.bioconductor.org/search/).

The source code of add-on packages from [GitHub](https://github.com/) can be
viewed directly on GitHub, or after downloading the code to your computer by
clicking the green `Code` button on the repository page (e.g.,
`https://github.com/JesseAlderliesten/checkrpkgs`), choosing `Download ZIP`, and
unzipping the downloaded file (right-click on the file and choose `extract all`).

### Basic method
The simplest way to get the source code of a function is to type the name of the
function, *without* the brackets. For example, to see what happens when using
`sd()` to calculate the standard deviation:

```{r}
sd
```

Some special cases:

- To distinguish functions with the same name from different packages, specify
  the package followed by two colons: `<pkg>::<func>`.
- For non-exported functions, specify the package followed by three colons:
  `<pkg>:::<func>` (using only two colons will result in the error
  `'<func>' is not an exported object from 'namespace:<pkg>'`; if that
  error appears when using three colons, you are probably looking in the wrong
  package, use `getAnywhere(<func>)` to check in which package `<func>` is
  defined). Non-exported functions should **not** be used in code because they
  might change.
- For operators such as `%in%` (see `help("match")` which start with a symbol,
  use backticks (`) around the name:
  
```{r}
`%in%`
```

### getAnywhere
A more robust alternative to the [basic method][Basic method] outlined above is
to use `getAnywhere("<func>")`, which looks in more places and finds
internal functions without the need to specify in which package a function is
defined. Although the quotes around the function name are not necessary when
looking for the source code of normal functions, they are required when looking
for the source code of operators such as `%in%`, such that it is most robust to
always use them.

#### UseMethod
If `getAnywhere("<func>")` returns `UseMethod("<func>")`, the function
has different methods for different object classes and is
[S3-generic](https://cran.r-project.org/doc/manuals/R-intro.html#Object-orientation).
First use `methods("<func>")` to get an overview of the available methods;
then get the source code of a particular method by using
`getAnywhere("<function.class>")`, where `<function.class>` is an item from the
overview that was returned by `methods("<func>")`. The advantage over simply
using `"<function.class>"` is that `getAnywhere("<function.class>")` also works
for functions that are not exported, which is indicated
in the overview of `methods(<func>)` by an asterisk and the remark
`Non-visible functions are asterisked`.

For example, `UseMethod("mean")` in the output of `getAnywhere("mean")`
indicates `mean` is an S3-generic:
```{r, eval = TRUE}
getAnywhere("mean")
```

First use `methods("mean")` to get an overview of the available methods:
```{r, eval = TRUE}
methods("mean")
```

The output shows, among others, the methods `mean.Date` and `mean.default`.
Then use `getAnywhere("mean.Date")` to see the source code of the method `mean`
used with objects of class `Date`. 

```{r, eval = TRUE}
getAnywhere("mean.Date")
```

Or use `getAnywhere("mean.default")` to see the source code of the method `mean`
used with objects of classes not listed in the output of `methods("mean")`:

```{r, eval = TRUE}
getAnywhere("mean.default")
```

#### standardGeneric
If `getAnywhere("<func>")` returns `standardGeneric("<func>")`, the
function has different S4-methods (see
`help("Introduction", package = "methods")`) for different object
classes. Use `showMethods("<func>")` to get an overview of the available
methods in all **attached** packages, or use `<pkg>:::<func>` to get an
overview of the available methods from package `<pkg>`. Finally, provide the
function name as argument `f` and the selected method as a character vector to
argument `signature` (see also `help("signature")`) of function `getMethod()` to
get the source code of a particular method:
`getMethod(f = "<func>", signature = c(target = "<class>", current = "<class>"))`.

The example below shows how to find the source code of method `cbind2()` used by
the `Matrix` package to combine two matrices that both have the `Matrix` class
as defined by package `Matrix`.

```{r, eval = TRUE}
library(Matrix)
getAnywhere("cbind2")

# standardGeneric("cbind2") indicating that cbind2() is an S4 function, so use
# showMethods("cbind2") to get an overview of the different methods:
showMethods("cbind2")

# To get the source code of 'cbind2' used with two matrices that both have the
# 'Matrix' class defined by package 'Matrix'
getMethod(f = "cbind2", signature = c(x = "Matrix", y = "Matrix"))
```

#### .Internal or .Primitive
If `getAnywhere("<func>")` returns `.Internal` or `.Primitive`, the function
is internal or primitive. The source code of such functions can be viewed at
code [repositories][Repositories], or on your computer if you have installed R
[from source](https://cran.r-project.org/sources.html) using Rtools (see the
section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
`vignette("install_r", package = "checkrpkgs")`).

Locate the file `src/main/names.c` and look in the first column ('printname')
for the name of the R function to find the appropriate c-entry which is given in
the second column of that file. Then either search for that c-entry using the
GitHub search, or manually locate the c-file (the name of the file is the
c-entry without the prefix `do_`) in `src/main` to get the file with the source
code.

For example, `getAnywhere("matrix")` shows, among others, the line
`.Internal(matrix(data, nrow, ncol, byrow, dimnames, missing(nrow), missing(ncol)))`.
The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `matrix` in the first column:
`{"matrix",	do_matrix,	0,	11,	7,	{PP_FUNCALL, PREC_FN,	0}}`
Searching `src/main` for `matrix` gives file
[array.c](https://github.com/r-devel/r-svn/blob/main/src/main/array.c) as one of
the results. That file contains the source code of `matrix`. Similarly,
`getAnywhere("log10")` shows `function (x)  .Primitive("log10")`. The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with `log10` in the first column:
`{"log10",	do_log1arg,	10,	1,	1,	{PP_FUNCALL, PREC_FN,	0}}`. Searching
`src/main` for `log1arg` gives file
[arithmetic.c](https://github.com/r-devel/r-svn/blob/main/src/main/arithmetic.c)
as one of its results. That file contains the source code of `log10`.

Sometimes an R function is defined in a file that defines multiple functions and
thus has a general name. Then the file with c-code will have a similar general
name. For example, the source code of `make.names()` is defined in
`src/library/base/R/character.R` which contains (among others) the code
`.Internal(make.names(names, allow_))`, and the c-code for `do_makenames()` is
in file `src/main/character.c`.

#### .Call
If `getAnywhere("<func>")` returns code that contains `.Call(...)`, the
function contains a call to C or C++ code. Although the file `src/<pkg>.h`
(e.g.,
[src/methods.h](https://github.com/r-devel/r-svn/blob/main/src/library/methods/src/methods.h)
if code is from package `methods`) contains an overview of the C or C++ code
included in a package, your best bet for finding the source code is
[searching](https://docs.github.com/en/search-github) the relevant part of the
[GitHub](https://github.com/r-devel/r-svn/) repository (e.g.,
`https://github.com/r-devel/r-svn/tree/main/src/library/methods/src` for code
from package `methods`), or installing the package from source (see the section
'Rtools' in the vignette *Installing R, Rtools and RStudio*:
`vignette("install_r", package = "checkrpkgs")`) and searching in the `src`
folder of the downloaded code.

## Documentation and help
- [Instructions](https://bioconductor.org/install/) on installing
  [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
  packages and [documentation](https://docs.r-universe.dev/bioconductor/) on the
  collaboration between BioConductor and the
  [R-universe](https://r-universe.dev/search)
- [Documentation](https://docs.r-universe.dev/) from the
  [R-universe](https://r-universe.dev/search)
- Search engines specific for R: [METACRAN](https://r-pkg.org/),
  [r-project](https://search.r-project.org/), [Rseek](https://www.rseek.org/),
  [R-universe](https://r-universe.dev/search)
- Section
  [Add-on packages](https://cran.r-project.org/doc/manuals/R-admin.html#Add_002don-packages)
  (which contains a section about installing packages) in the
  [R Installation and Administration manual](https://cran.r-project.org/doc/manuals/r-release/R-admin.html)
  manual
- Section 'Documentation and help' in the vignette
  *Installing R, Rtools and RStudio*:  
  `vignette("install_r", package = "checkrpkgs")`
- Section [Troubleshooting] above
- The book [What They Forgot to Teach You About R](https://rstats.wtf/) by J.
  Bryan, J. Hester, S. Pileggi, and E. D. Aja
