---
title: "Instructions about R packages"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Instructions about R packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction and notation
This vignette contains information on installing and getting information about R
packages. It also explains how to obtain the source code of R functions.

In this vignette, `<pkgname>` is used to refer to a package name that should be
filled in to get working code, e.g., `<pkgname>` should be replaced with `grid`
if you want to obtain information about package `grid`, and with `stats` if you
want to obtain information about package `stats`. Similarly, `<funcname>` is
used to refer to a function name that should be filled in to get working code.


## R packages
A [package](https://cran.r-project.org/doc/FAQ/R-FAQ.html#What-is-the-difference-between-package-and-library_003f)
is a standardized collection of material extending R by providing code, data, or
documentation. You have to install a package (see section 'Installing packages'
below) and then need to run `library(<pkgname>)` to be able to use the functions
of that package.

The
[base](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-in-R)
packages are always installed together with R. The
[recommended](https://cran.r-project.org/doc/FAQ/R-FAQ.html#Add_002don-packages-from-CRAN)
packages are installed with binary distributions of R. Together, the base and
recommended packages are the 'high priority' packages. Since R 4.4.0,
`tools::standard_package_names()` contains a list with the names of these
packages. To see which high-priority packages are currently installed, run
`sort(unname(installed.packages(priority = "high")[, "Package"]))`. Package
'translations' is not a recommended package, but will be installed with R if
that option is set during the installation.

### Installing packages
Run R (or RStudio) as administrator: right-click on the R (RStudio) icon and
select 'Run as administrator'. Remember that after installing a package, you
need to run `library(<pkgname>)` to be able to use the functions of that
package.

Installing an old version of a package might require installing Rtools to build
the packages from source, see the section 'Rtools' in the vignette
*Installing R, Rtools and RStudio*: `vignette("install_r", package = "checkrpkgs")`.

#### CRAN
[CRAN](https://cran.r-project.org/) or any of its
[mirrors](https://cran.r-project.org/mirrors.html) (i.e., websites hosted in
various parts of the world with the same content as the main CRAN website) host
R [packages](https://cran.r-project.org/web/packages/index.html).

Thematic collections are organised in
[task views](https://cran.r-project.org/web/views/). To install all core packages
of a task view, install package [ctv](https://CRAN.R-project.org/package=ctv)
and run `ctv::install.views("<taskviewname>", coreOnly = TRUE)`. To update these
packages, use `ctv::update.views("<taskviewname>", coreOnly = TRUE)`.

CRAN packages that have been recently archived are available at
[CRANhaven](https://www.cranhaven.org/);

The code below can be used to install packages from 
[CRAN](https://cran.r-project.org/web/packages/index.html):

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
bool_install <- !vapply(X = pkgs_new, FUN = requireNamespace,
                        FUN.VALUE = logical(1), quietly = TRUE)
pkgs_install <- pkgs_new[bool_install]
if(length(pkgs_install) > 0L) {
  install.packages(pkgs = pkgs_install, lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
```

The code below can be used to install an old version of a package from CRAN,
using package [remotes](https://CRAN.R-project.org/package=remotes):

```{r, eval = FALSE}
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_version(package = <pkgname>, version = "1.23", dependencies = NA,
                         upgrade = "ask", quiet = FALSE, build_vignettes = TRUE,
                         verbose = getOption("verbose"))
```

Alternatively, use the URLs pointing to older versions at the installation page
of CRAN packages (at `Downloads` > `Old sources` > `<pkgname> archive`) to
install older versions using base R. For example, to install an older version of
package [deSolve](https://CRAN.R-project.org/package=deSolve):
```{r, eval = FALSE}
install.packages(pkgs = "https://cran.r-project.org/src/contrib/Archive/deSolve/deSolve_1.40.tar.gz",
                 lib = .libPaths(), repos = NULL, dependencies = NA,
                 type = "source", verbose = getOption("verbose"),
                 quiet = FALSE)
```

#### BioConductor
[Bioconductor](https://bioconductor.org/) has many
[packages](https://bioconductor.org/packages/release/BiocViews.html#___Software),
with thematic
[collections](https://bioconductor.org/packages/release/BiocViews.html).

The code below can be used to install packages from
[CRAN](https://cran.r-project.org/web/packages/index.html) or
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html#___Software), using
package [BiocManager](https://CRAN.R-project.org/package=BiocManager). The
packages are from a specific BioConductor release (here version 3.22), which are
tied to a
[specific version](http://bioconductor.org/about/release-announcements/#release-versions)
of R to make sure that packages can be installed together.

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = pkgs_new, lib = .libPaths(), dependencies = NA,
                     type = getOption("pkgType"), verbose = getOption("verbose"),
                     update = FALSE, ask = TRUE, checkBuilt = TRUE,
                     force = FALSE, version = "3.22")
```

Older versions of Bioconductor packages can be
installed by changing the value of argument `version`, see the
[overview](https://bioconductor.org/about/release-announcements/) of
Bioconductor versions with the corresponding R version. Information on these
older versions can be found by visiting the appropriate version of Bioconductor,
e.g., `https://bioconductor.org/packages/3.21/BiocViews.html`.

#### Github
Many R packages are available from [GitHub](https://github.com/). The code below
can be used to install GitHub packages, using package
[remotes](https://CRAN.R-project.org/package=remotes).

```{r, eval = FALSE}
pkgs_new <- c(<pkgname>, <pkgname>)
if(!requireNamespace("remotes", quietly = TRUE)) {
  install.packages(pkgs = "remotes", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
remotes::install_github(repo = pkgs_new, dependencies = NA, upgrade = FALSE,
                        force = FALSE, quiet = FALSE, build_vignettes = TRUE,
                        lib = .libPaths(), verbose = getOption("verbose"))
```

#### Other repositories

- [Neuroconductor](https://neuroconductor.org/list-packages/all);
- [R-Forge](https://r-forge.r-project.org/) with a 
  [GitHub mirror](https://github.com/r-forge) and thematic
  [collections](https://r-forge.r-project.org/softwaremap/trove_list.php);
- [R-multiverse](https://r-multiverse.org/overview.html) with
  thematic [collections](https://r-multiverse.org/topics/);
- [rOpenSci](https://ropensci.org/packages/all/) with
  thematic [collections](https://ropensci.org/packages/);
- [R universe](https://r-universe.dev/search) with an overview of
  [datasets](https://r-universe.dev/datasets) included in R packages.


### Updating packages
To get the version number of an installed package, run
`utils::packageVersion("<packagename>")`. For R packages available at CRAN,
versions can be compared using [diffify](https://diffify.com/R) and a
chronological overview of changes is available at
[CRANberries](https://dirk.eddelbuettel.com/cranberries/).

Updating out-of-date packages prevents compatibility issues between
already-installed and newly-installed packages. However, it *changes* the
version of already-installed packages, which might be unwanted.

The following code can be used to install the latest version of packages from
[CRAN](https://cran.r-project.org/web/packages/index.html):

```{r, eval = FALSE}
utils::update.packages(lib.loc = .libPaths(), ask = TRUE, dependencies = NA,
                       verbose = getOption("verbose"), quiet = FALSE,
                       checkBuilt = TRUE, type = getOption("pkgType"))
```

Alternatively, the following code can be used to update packages to a specific
BioConductor release (here version 3.22), see section 'Installing packages >
BioConductor' above:

```{r, eval = FALSE}
if(!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages(pkgs = "BiocManager", lib = .libPaths(), dependencies = NA,
                   type = getOption("pkgType"), verbose = getOption("verbose"),
                   quiet = FALSE)
}
BiocManager::install(pkgs = character(), lib = .libPaths(), verbose = FALSE,
                     update = TRUE, ask = TRUE, checkBuilt = TRUE,force = FALSE,
                     version = "3.22")
```

### Troubleshooting
- If a package appears not to be installed when you want to use a function from
  it, remember you need to run `library(<pkgname>)` to be able to use a function
  from package `<pkgname>`.

- If a package is not functional, re-install it using argument `force = TRUE`.
  You can also use `checkrpkgs::get_details_pkgs(pkgs = <pkgname>)` to check
  which dependencies and other system requirements it has and then use
  `checkrpkgs::find_nonfunc_pkgs(pkgs = <pkgnames>)` to check if all the
  dependencies are installed and functional.

- If the warning `lib = <pkgname> is not writeable` occurs, you probably forgot
  to run R (or RStudio) as administrator. Close R (RStudio) and right-click on
  the R or RStudio icon and select `Run as administrator` to restart it as
  administrator.

- If the question
  `Do you want to install from sources the packages which need compilation?` is
  asked, possibly accompanied by the remark
  `There are binary versions available but the source versions are later`, you
  should choose `No` if you have *not* installed
  [Rtools](https://cran.r-project.org/bin/windows/Rtools/): you will then get
  slightly less up to date package versions but a faster installation.
  
  If you want the latest version without using Rtools, you can try again a few
  days later: it usually takes a bit longer for the CRAN binaries (i.e., the
  package versions that do not have to be build from source) to be updated than
  for the source versions used by Rtools.
  
  If you have installed Rtools you can choose `Yes` to install the latest
  package versions by building them from source. To install Rtools, see the
  section 'Rtools' in the vignette *Installing R, Rtools and RStudio*:
  `vignette("install_r", package = "checkrpkgs")`.

- If the warning `package <pkgname> was built under R version 'x.y.z'` occurs,
  you installed a binary package (i.e., not by building from source from source)
  that was compiled for an earlier version of R than the version of R you are
  currently using (see `R.version.string` for your current R version). The
  warning is issued because packages are not tested on versions of R that are
  older than the version they were built on. Therefore it is best to update R
  when installing packages.

- The warning `package '<pkgname>' is not available (for R version x.y.z)` can
  have a lot of reasons. Double-check the package name (which is case-sensitive).
  See the overview in this
  [stackoverflow answer](https://stackoverflow.com/a/25721890/32365738) for
  other possible reasons.

- If errors occur when loading packages that require Java, make sure the 64-bit
  [version of Java](https://www.java.com/download/manual.jsp) is installed on
  64-bit PCs.


## Information about packages

### Not-yet-installed packages
The source code of base R packages can be obtained from
[GitHub](https://github.com/r-devel/r-svn/) (see the section 'Getting the source
code' below), and the installation pages of packages on CRAN frequently contain
links to the GitHub pages where their source code can be viewed.

A PDF-file with all the help files of the standard and recommended packages is
[available](https://cran.r-project.org/doc/manuals/r-release/fullrefman.pdf).
The manuals and help pages of all CRAN packages can be
searched [online](https://search.r-project.org/) or from within R using
`utils::RSiteSearch()`. `utils::help.search()` searches uses fuzzy matching or
regular expressions while searching the help system, which can be disabled by
setting argument `agrep` to `FALSE` to search faster and return fewer results:
`utils::help.search(..., agrep = FALSE)`.

`tools::CRAN_package_db()` gives information about packages available from
[CRAN](https://cran.r-project.org/web/packages/index.html);
`BiocManager::available()` gives the names of packages available from
[BioConductor](https://bioconductor.org/packages/release/BiocViews.html).

#### Package dependencies
Use `tools::package_dependencies(packages = "<pkgname>", recursive = TRUE)` to
see dependencies and `tools::dependsOnPkgs(pkgs = "<pkgname>", recursive = TRUE)`
to see reverse dependencies.

### Already-installed packages
The locations where packages have been installed can be obtained with
`.libPaths()`, and the names of all installed packages can be obtained with
`list.files(path = .libPaths(), recursive = FALSE)`. The repository from which a
package was installed can be obtained from the `Repository` and `URL` fields of
package descriptions: `installed.packages(fields = c("Repository", "URL"))`.
This output can be used to re-install the packages after installing a new
version of R.

Information on a package and its functions is available from within R after the
package has been installed and loaded (i.e., `library(<pkgname>)` has been run):

- Help page of a package: `help(topic = "<pkgname>")`.
- Package functions: `ls(getNamespace("<pkgname>"))` gives their names as a
  character vector; `help(package = "<pkgname>")` gives an overview with links
  to their help-pages (does not work if there is no file `<pkgname>.R` in folder
  `\R`).
- Vignettes: list them with `vignette(package = "<pkgname>")`; or show them in a
  browser: `browseVignettes(package = "<pkgname>")`.
- Show installed package version: `utils::packageVersion("<pkgname>")`.

Information about functions, methods, and classes can also be obtained (see also
the section 'Getting the source code' below):

- Help page of a function: `help("<funcname>")`.
  to distinguish between functions from different packages:
  `help("<funcname>", package = "<pkgname>")`; use quotes to obtain the help
  page of an operator: `help("%in%")`.
- List of function arguments: `args("<funcname>")`.
- List of methods that have been defined for a generic class:
  `utils::methods(class = "classname")`;
  `attr(methods(class = "classname"), "info")` for S3-methods;
  `methods::showMethods(classes = "classname", where = getNamespace("<pkgname>"))`
  for S4-methods.
- List of methods that have been defined for a generic function:
  `utils::methods("<funcname>")`.
- List of objects (including functions) whose name contains a certain string:
  `utils::apropos("<funcname>")`.

Information about the citation of packages is available through
`utils::citation("<pkgname>")`, with `utils::citation()` to cite R itself.

### Which packages are used?
The function `loadedNamespaces()` can be used to see which packages are loaded.
To see which packages are used in a script, look for `library`, `require`,
`namespace` (e.g., `loadNamespace`, `requireNamespace`), `::`, and `:::`.

To see which packages are mentioned in comments, also look for:

- `conductor`: [Bioconductor](https://bioconductor.org/),
  [Neuroconductor](https://neuroconductor.org/)
- `CRAN`: [CRAN](https://cran.r-project.org/)
- `forge`: [R-Forge](https://r-forge.r-project.org/)
- `Git`: [GitHub](https://github.com/)
- `package`
- `rOpenSci`: [rOpenSci](https://ropensci.org/packages/)
- `verse`: [R-multiverse](https://r-multiverse.org/),
  [R-universe](https://r-universe.dev/packages),
  [tidyverse](https://tidyverse.org/packages/), etc

## Getting the source code

### Acknowledgement
Partly based on:

- Bryan J. 2015. [Accessing R Source](https://github.com/jennybc/access-r-source).
- Ligges U. 2006. Accessing the sources.
[RNews 6(4):43-45](https://journal.r-project.org/articles/RN-2006-035/).
- A post from [StackOverflow](https://stackoverflow.com/a/19226817).

### Repositories
The source code of the base R packages, for the current as well as earlier
versions of R, is available at [CRAN](https://cran.r-project.org/src/base/)
and at the [SVN-project](https://svn.r-project.org/R/). Searching the source
code is easiest using the [GitHub mirror](https://github.com/r-devel/r-svn/) of
the latter.

The source of the add-on packages from
[CRAN](https://cran.r-project.org/web/packages/index.html)
is available at [METACRAN](https://github.com/cran), an *unofficial* CRAN mirror.
The source code of [BioConductor](https://bioconductor.org/) packages is
available [here](https://code.bioconductor.org/search/).

### Basic method
The simplest way to obtain the source code of a function is typing the name of
the function, *without* the brackets. E.g., `sd` returns
`function (x, na.rm = FALSE) sqrt(var(if (is.vector(x) || is.factor(x)) x else as.double(x), na.rm = na.rm))`.

Some special cases:

- For operators such as `%in%` which start with a symbol, use backticks around
  the name.
- To differentiate functions with the same name defined in different packages,
  specify the package followed by two colons: `<pkgname>::<funcname>`.
- For non-exported (i.e., internal) functions, specify the package followed by
  three colons: `<pkgname>:::<funcname>`. Non-exported functions should *not* be
  used directly: they might change.

### getAnywhere
A more robust alternative to the basic method outlined above is using
`getAnywhere("<funcname>")`: `getAnywhere()`
looks in more places and finds internal functions without the need to specify
in which package a function is defined. Although the quotes around the function
name are not necessary when looking for normal functions, they are required when
looking for the source code of operators such as `%in%`, such that it is most
robust to always use them.

#### UseMethod
If `getAnywhere("<funcname>")` returns `UseMethod("<funcname>")`, the function
has different methods for different object classes and is
[S3-generic](https://cran.r-project.org/doc/manuals/R-intro.html#Object-orientation).
Use `methods("<funcname>")` to get an overview of the available methods and then
`getAnywhere("<function.class>")`, where `<function.class>` is an item from the
overview returned by `methods("<funcname>")`, to obtain the source code of a
particular method.

For example, `getAnywhere("mean")` returns
`function (x, ...) UseMethod("mean")`, indicating it is an S3-generic. Using
`methods("mean")` returns `mean.Date`, `mean.default`, and other methods.
`getAnywhere("mean.Date")` returns the source code for the method used with
objects of class 'Date', and `getAnywhere("mean.default")` returns the source
code for the default method (i.e., the method used with objects of classes not
listed in the output of `methods("mean")`):

```{r, eval = FALSE}
function (x, trim = 0, na.rm = FALSE, ...) {
    if (!is.numeric(x) && !is.complex(x) && !is.logical(x)) {
        warning("argument is not numeric or logical: returning NA")
        return(NA_real_)
    }
    if (isTRUE(na.rm))
      x <- x[!is.na(x)]
      < omitted further output >
```

If the overview of
`methods(<funcname>)` includes the remark `Non-visible functions are asterisked`,
it means that these functions are not exported and you need to use
`getAnywhere("<method.class>")` or `<pkgname>:::<method.class>`.

#### standardGeneric
If `getAnywhere("<funcname>")` returns `standardGeneric("<funcname>")`, the
function has different S4-methods (see `methods::Introduction`) for different
object classes. Use `showMethods("<funcname>")` to get an overview of the
available methods in all loaded packages, or `<pkgname>:::<funcname>` to get an
overview of the available methods from package `<pkgname>`. Then provide the
function name as argument `f` and the *entire line* of the selected method as
argument `signature` to function `getMethod()`, i.e.,
`getMethod(f = "<funcname>", signature = c(target = "<classname>", current = "<classname>"))`
to obtain the source code of a particular method.

For example, if the Matrix package is installed and loaded (i.e.,
`library(Matrix)` has been run), `getAnywhere("all.equal")` returns
matches for package Matrix and for package base. Selecting the match from
package Matrix (i.e., `getAnywhere("all.equal")[1]`) returns, among others,
`standardGeneric("all.equal")` indicating that 'all.equal' is an S4 function.
`showMethods("all.equal")` gives an overview of the different methods. Then
providing the function name as argument `f` and the *entire line* of the
selected method as argument `signature` to function `getMethod()`:
`getMethod(f = "all.equal", signature = c(target = "vector", current = "Matrix"))`
returns the source code for that particular method:

```{r, eval = FALSE}
function (target, current, ...) {
    .local <- function (target, current, check.type = check.class, 
        check.class = TRUE, check.attributes = TRUE, check.factors = FALSE, 
        ...) {
        msg <- .M.attr.all.equal(target, current, check.type = check.type, 
            check.class = check.class, check.attributes = check.attributes, 
            check.factors = check.factors, ...)
        < omitted further output >
```

#### .Internal or .Primitive
If `getAnywhere("<funcname>")` returns `.Internal` or `.Primitive`, the function
is internal or primitive. The source code of such functions can be viewed at the
code repositories such as [GitHub](https://github.com/r-devel/r-svn/), or on
your computer if you have installed R
[from source](https://cran.r-project.org/sources.html).

Locate the file 'src/main/names.c' and look in the first column for the name of
the R function to find the appropriate c-entry which is given in the second
column of that file. Then either search for that c-entry using the GitHub
search, or manually locate the c-file (the name of the file is the c-entry
without the prefix 'do_') in 'src/main' to get the file with the source code.

For example, `getAnywhere("matrix")` shows, among others, the line
`.Internal(matrix(data, nrow, ncol, byrow, dimnames, missing(nrow), missing(ncol)))`.
The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with 'matrix' in the first column:
`{"matrix",	do_matrix,	0,	11,	7,	{PP_FUNCALL, PREC_FN,	0}}`
Searching 'src/main' for 'matrix' gives file
[array.c](https://github.com/r-devel/r-svn/blob/main/src/main/array.c) as one of
the results. That file contains the source code of 'matrix'. Similarly,
`getAnywhere("log10")` shows `function (x)  .Primitive("log10")`. The file
[src/main/names.c](https://github.com/r-devel/r-svn/blob/main/src/main/names.c)
has as entry with 'log10' in the first column:
`{"log10",	do_log1arg,	10,	1,	1,	{PP_FUNCALL, PREC_FN,	0}}`. Searching
'src/main' for 'log1arg' gives file
[arithmetic.c](https://github.com/r-devel/r-svn/blob/main/src/main/arithmetic.c)
as one of its results. That file contains the source code of `log10`.

#### .Call
If `getAnywhere("<funcname>")` returns code that contains `.Call(...)`, the
function contains a call to C or C++ code. Although the file `src/<pkgname>.h`
(e.g.,
[src/stats.h](https://github.com/r-devel/r-svn/blob/main/src/library/stats/src/stats.h)
if code is from package `stats`) contains an overview of
the C or C++ code included in a package, your best bet for finding the source
code is using the [GitHub](https://github.com/r-devel/r-svn/) search, or
installing the package from source and
searching in the `src` folder of the downloaded code.


## Documentation and help
- [Instructions](https://bioconductor.org/install/) on installing
  [BioConductor](https://bioconductor.org/packages/release/BiocViews.html)
  packages
- [Search engines](https://cran.r-project.org/search.html) specific for R
- Section
  [Add-on packages](https://cran.r-project.org/doc/manuals/R-admin.html#Add_002don-packages)
  in the
  [R Installation and Administration manual](https://cran.r-project.org/doc/manuals/r-release/R-admin.html)
  manual
- Section 'Documentation and help' in the vignette
  *Installing R, Rtools and RStudio*:  
  `vignette("install_r", package = "checkrpkgs")`
- Section `R packages` > `Troubleshooting` above
- The help-page `?checkrpkgs::find_nonfunc_pkgs`
